# Data and data handling (Week 5) {#sec-05-prep2}

A clinic spreadsheet kept by hand for years has a title sitting in the first row, a blank row under the headers, dates written three different ways, and a column where someone typed 'n/a' one month and '-' the next. Nobody planned it that way. It happened one edit at a time, and files like it are what real analysis starts from. The dataset we typed out in Chapter 1 was clean because we invented it. This chapter deals with the other kind.

Making a file like that usable is most of what data analysis is, and none of it counts unless it's written down: a cleaned dataset nobody can reproduce is one person's private result. So the chapter runs from the messy spreadsheet all the way through to a saved script that rebuilds the lot from nothing. In between sits the step that catches almost everyone out the first time, which is telling R where a file lives on our own computer, getting it in, and checking what arrived instead of assuming. Week 6 is reading week; Chapter 3 picks up from here.

## Chapter intended learning outcomes

By the end of this week you will be able to:

1.  Identify features of a spreadsheet that will prevent it being analysed, and restructure it
2.  Import a CSV into R using `readr::read_csv()` and a correctly specified path built with `here::here()`, and diagnose and resolve the most common import failures
3.  Examine an imported dataset's dimensions, variable names, and types with `dim()`, `names()`, and `dplyr::glimpse()`, and its missing values with `is.na()`
4.  Write, save, and re-run an R script that imports data, summarises it, and produces a plot
5.  Locate a downloaded file on your own operating system, construct or copy a valid path to it, and explain what a CSV file physically contains

## Setup

There's nothing to read in yet, so there's nothing to set up in this chapter.

```{r}
#| label: setup-hidden
#| include: false
options(width = 76)
```

## Tidy data, in a spreadsheet

### What makes data usable

A dataset R can work with follows three simple rules: one variable per column, one observation per row, one value per cell. Data laid out this way is called **tidy**. It sounds obvious written down, but very few spreadsheets we get are tidy. People merge cells to make a heading look nice. They use colour instead of writing a word down. They leave blank rows for visual spacing. None of that survives import, and some of it actively breaks it.

### What breaks an import

Picture reading the sheet in the screenshot below the way a person would, working down it row by row, and watching where it stops behaving like one clean table.

Row 1 is a title, 'Clinic Visit Log - Week 12', stretched across six columns as one merged cell. A heading merged like that looks fine to a person. But when R reads the underlying columns back out, most of them come back empty, because the merge only stored the value in one of the cells it covers. Row 2 does the same thing on a smaller scale, merging 'Measurements' across just the Weight and Height columns, so a proper header for either one only turns up in row 3.

Read on down the Weight and Height columns themselves and one row breaks the pattern. Most patients are plain numbers, kilograms and metres. One row is text instead: `195 lbs` and `6 ft`. R reads a whole column as text the moment a single cell in it isn't a bare number. This is the same problem Chapter 1's `round()` example was solving for us by hand, turning a raw scale reading into a clean whole-kilogram number before it ever went in the log. Nobody has done that here, and R has no way to do it on our behalf.

Off to the side, in a cell with no column of its own, sits a note about one specific patient's missed follow-up, disconnected from the table it's about. Further down, after a blank row inserted purely for visual spacing, which becomes a row of missing values, `NA`, in every column the moment it's imported, sits a second, smaller table: a follow-up schedule, unrelated to the patient log above it. R expects one table per sheet, so this second table gets read as more rows of the first one, and the result makes no sense.

**Spreadsheet exercise (ILO 1).** `data/05-prep2-messy-clinic.xlsx` is a small, made-up clinic visit log, built specifically to have every one of these problems at once. Open it in Excel, or whatever spreadsheet software you have.

![`05-prep2-messy-clinic.xlsx` open in a spreadsheet application: a merged title row, a merged 'Measurements' sub-header spanning the Weight and Height columns, a block of rows highlighted yellow with a caption below the table explaining what the highlighting means, a stray note in one row about a patient's missed follow-up, a blank spacer row, and a second, separate table further down the sheet.](../images/05-prep2-messy-spreadsheet.png){fig-alt="A spreadsheet showing a clinic visit log with a merged title row, a merged sub-header over the Weight and Height columns, several rows highlighted yellow with a caption below reading 'Highlighted are smokers', a note in one row about a missed follow-up, a blank row, and a second table for a follow-up schedule." width='90%' fig-align='center'}

Go through the sheet and list every feature above that you can find in it. There is also one problem in this sheet that wasn't mentioned above. Try to build a clean version based on the principles above. Save your clean version as a `.csv` file. The Solutions section describes the fixes.

## Path skills: Finding a file, and telling R where it is

We just saved a clean CSV in the exercise above. Before we can read it, or anything else, into R, we need to know exactly where the file we want is located. The location of a file is referred to at the file path, or just path. This section covers the path finding skills you'll use for the rest of the course. Just like other R skills, we're using a reproducible, text based method instead of pointing and clicking on folders.

**Where downloads go.** When you download a file from a browser, most browsers save it into a folder called `Downloads` by default, without asking. On Windows that's usually `C:\Users\yourname\Downloads`; on a Mac it's usually `/Users/yourname/Downloads`. If you can't find a file you just downloaded, that folder is the first place to check. Confirm where your own browser is set to save to, since some let you change it.

**Reading a path.** A file path is a set of directions from somewhere fixed down to one specific file, one folder at a time, separated by slashes. `data/mph_nhanes_20241107.csv` reads as 'the folder called `data`, then inside it the file called `mph_nhanes_20241107.csv`.' A longer path has more folders to walk through before reaching the file.

**Copying a path.** We'll rarely type a full path out by hand; instead, we copy it straight from the file browser.

-   **Windows:** hold Shift and right-click the file, then choose **Copy as path**.
-   **macOS:** right-click the file, then hold the **Option** key, which changes the menu, and choose **Copy 'filename' as Pathname**.

**Backslashes versus forward slashes.** A path copied this way on Windows uses backslashes, `\`, for example `C:\Users\yourname\Documents\ies_mph\data\file.csv`. R uses forward slashes, `/`, instead, and treats a lone backslash as the start of a special character, not a folder separator. So a pasted Windows path usually needs every `\` replaced with `/` before R will accept it.

**What a CSV is.** Open a `.csv` file in a plain text editor, Notepad on Windows or TextEdit on a Mac, not in spreadsheet software, and you'll see plain text: one line per row, values separated by commas. There's no hidden formatting behind it, which is why it's such a safe format to hand between different pieces of software, R included.

## Getting data in

### `read_csv()` and `here()`

`readr`'s `read_csv()` reads a `.csv` file into R as a data frame. It needs a path telling it where that file is, using the path skills from earlier in this chapter. Building that path with `here()`, from the `here` package, means the same code works on any machine, regardless of where the project folder sits on that machine's disk: `here()` starts from the project folder and adds the pieces you give it, one folder at a time.

This is a different function from `read.csv()`, base R's own version, which Chapter 1 used deliberately, and from `write.csv()`, which Chapter 1 used to save `visits` out. Base R's versions weren't wrong there, only the simpler tool for four hand-typed patients. From here on we use `readr`'s `read_csv()`, and later `write_csv()`, instead. They read faster on a file this size. They also don't silently guess at things the way base R's older functions do: `read.csv()` used to turn every text column into a factor by default unless told not to, a behaviour that caught out R users for years. `read_csv()` and `write_csv()` are the pair this book, and the wider tidyverse, are built around.

This chapter needs three packages: `readr` for reading files, `here` for building paths, and `ggplot2` for the plot at the end. As Chapter 1 explained, installing a package and loading it are two different actions. Unless these are installed already, run this once in the console, not inside this document:

``` r
install.packages(c('readr', 'here', 'ggplot2'))
```

After that, `library()` is all we need, in this session and every session after it:

```{r}
#| label: load-packages
#| message: false
library(readr)
library(here)
```

`library(here)` prints one line when it loads, which this book hides because it's too wide to fit the page. In your own console it says `here() starts at ...` followed by the full path to your project folder. Read it the first time you see it, because it's `here()` telling you which folder it will build every path from. If that path isn't the project folder you built in Chapter 1, nothing below will find its files, and this line is how you'd find that out.

### Four things that go wrong

Almost every import problem we'll see is one of four things.

**A wrong filename.** Close, but not exact:

```{r}
#| label: err-wrong-path
#| error: true
read_csv(here('data', 'mph_nhanes_20241107_data.csv'))
```

The real file is called `mph_nhanes_20241107.csv`, no `_data` in the name. R's message states plainly that the file doesn't exist at the path it tried, which is what Chapter 1's 'file does not exist' error type looks like with a real file behind it.

```{r}
#| label: fix-wrong-path
#| eval: false
read_csv(here('data', 'mph_nhanes_20241107.csv'))
```

**A path pasted straight from Windows Explorer.** Shift+right-click → *Copy as path*, from earlier in this chapter, gives you a full path using backslashes, for example `C:\Users\yourname\Documents\new_folder\mph_nhanes_20241107.csv`. Paste that directly into R as a string, without changing anything:

```{r}
#| label: err-backslash
#| error: true
read_csv('C:\Users\yourname\Documents\new_folder\mph_nhanes_20241107.csv')
```

This isn't even a 'file not found' error. It never gets that far. `\U` is one of a handful of backslash sequences R's parser treats specially inside a string, and `\Users` isn't a valid one, so R refuses to parse the string at all. Every Windows path copied this way starts with `C:\Users\...`, so this is the usual case. The fix from earlier in this chapter still applies: replace every backslash with a forward slash before using a path in R. Better still, once the path is inside the project folder, `here()` avoids the problem entirely, since we never type the backslash-filled part of the path at all.

**The right file, wrong location.** Forgetting which subfolder the file lives in:

```{r}
#| label: err-file-not-there
#| error: true
read_csv(here('mph_nhanes_20241107.csv'))
```

The filename is right this time. What's missing is `'data'`, the subfolder it lives in. R's error message looks identical to the wrong-filename case above: the message tells us *that* nothing was found at that path, not *why* the path was wrong. We have to check that ourselves.

```{r}
#| label: fix-file-not-there
#| eval: false
read_csv(here('data', 'mph_nhanes_20241107.csv'))
```

**The wrong file type.** `data/05-prep2-messy-clinic.xlsx`, our own clinic file from the exercise above, is an Excel file, not a CSV. Reading it with `read_csv()` doesn't throw an error at all.

To see what came in, we need three functions. `dim()` reports how many rows and columns a dataset has, and `names()` lists its column names. Both get a proper introduction in the next section, where we use them on real data; here they're just the quickest way to look. `substr()`, from base R, takes a piece of text and keeps only part of it, given a starting and an ending position: `substr(x, 1, 50)` keeps the first 50 characters and throws the rest away. We need it because the column name that comes back is far too long to print in full.

There's one extra argument in the call below, `show_col_types = FALSE`. `read_csv()` normally prints a short report of what it found and what type it guessed for each column. We meet that report properly in a moment, on the real data file, where it's useful. Here it would just be several lines of nonsense about a file that hasn't imported, so we turn it off to keep the broken result below the only thing on screen.

```{r}
#| label: wrong-filetype
messy_attempt <- read_csv(here('data', '05-prep2-messy-clinic.xlsx'),
  show_col_types = FALSE)
dim(messy_attempt)
substr(names(messy_attempt), 1, 50)
```

No error, but nothing usable either: one column, zero rows, and a column name that's a fragment of raw XML. An `.xlsx` file is a small zip archive of several XML files bundled together. `read_csv()` has gone looking inside that archive and started reading one of those XML files as if it were plain text. This is the quietest kind of import failure there is: nothing crashes, nothing complains, and the result still looks like a data frame at a glance. Always check what came in, which is what the next section is for.

Put right, all four attempts above collapse into the one call we want, and this is the real import we use for the rest of this book. The file holds `nhanes_mph`, an extract of a large American health survey in which adults were interviewed and then given a physical examination, tens of thousands of them, not four. Chapter 3 introduces it properly:

```{r}
#| label: read-nhanes
#| message: false
nhanes <- read_csv(here('data', 'mph_nhanes_20241107.csv'))
```

We don't necessarily need to use the `here` package. We can use an absolute or a relative path instead, as below. However, if the RStudio project is set up as recommended, using `here` or a relative path is easier to work with.

```{r}
#| label: read-nhanes-rel-path
#| eval: false
nhanes <- read_csv('./data/mph_nhanes_20241107.csv')
```

```{r}
#| label: read-nhanes-abs-path
#| eval: false
nhanes <- read_csv('C:/Users/yourname/Documents/ies_mph/data/mph_nhanes_20241107.csv')
```

An absolute path like that one only works on the machine it was written on, since it names a specific person's folder by name. `here()` avoids that problem. It finds the project folder wherever it happens to sit. That's why this book uses it throughout, and why the restart test later in this chapter only works if our script uses `here()` instead of a path like the one above.

### The block of output we'll see

Run that line and `read_csv()` prints eight lines the moment it finishes. This book hides them, because one of them is too wide to fit the page, so here they are as they'll appear in your console:

```         
Rows: 41765 Columns: 27
── Column specification ────────────────────────────────
Delimiter: ','
chr  (3): smoking, phq.c, lcod
dbl (24): SEQN, age, gender, eth, edu, alc_wk, vpa_w, ...

ℹ Use `spec()` to retrieve the full column specification
  for this data.
ℹ Specify the column types or set `show_col_types = FALSE`
  to quiet this message.
```

It looks like something went wrong. Nothing did. This is a **message**, not a warning and not an error, and it appears every single time `read_csv()` reads a file successfully. Read it once, properly, because it's the first thing we know about a dataset we've never opened.

`Rows: 41765 Columns: 27` is the size of what came in. That's the same fact `dim()` gives us below, arriving before we ask for it. If a file we expected to have a few hundred rows reports three, we've learned that here instead of ten minutes later.

`Delimiter: ','` is `read_csv()` confirming it split each line at commas, which is what the 'comma' in 'comma separated values' means, and what we saw for ourselves earlier in this chapter when we opened a CSV in a text editor.

Then the column specification: `chr (3)` lists three columns read as character, that is text, and `dbl (24)` lists twenty-four read as numbers. `dbl` is short for 'double', which is R's ordinary way of storing a number that can have decimal places. This is `read_csv()` telling us what it *guessed*, one column at a time, and that guess is the thing worth checking. Notice `smoking` and `phq.c` sitting in the character list, and `gender` and `eth` sitting in the number list, and hold onto that: it comes back in the next section.

The two lines beginning with an `ℹ` are `read_csv()` offering a way to see the full specification, and a way to make the whole message go away. That second one is `show_col_types = FALSE`, the argument we used on the broken `.xlsx` import above. Now we know what it silences. Use it when we already know what's in a file and don't want the report again, not to make untidy output go away. 'Untidy output' and 'the only warning we were going to get' often look identical.

### A first look

The message already told us the size and the column types. Five functions let us check that against what we have.

```{r}
#| label: inspect-dim-names
dim(nhanes)
names(nhanes)
```

`dim()` gives rows, then columns. `names()` lists every column name, in order, useful for checking a name before we type it.

```{r}
#| label: inspect-head
head(nhanes)
```

`head()` prints the first six rows, so we can eyeball what real values look like. With 27 columns, most of them don't fit across the screen, and the printout says how many it left out.

```{r}
#| label: inspect-glimpse
dplyr::glimpse(nhanes)
```

`glimpse()`, from `dplyr`, turns the data on its side: one line per column, with its type and its first few values. That fits a wide dataset like this one on screen, every column included. We haven't loaded `dplyr` in this chapter, so we call it with the `package::function()` form from Chapter 1.

```{r}
#| label: inspect-summary
summary(nhanes$age)
```

`summary()`, run on one column, gives its minimum, maximum, quartiles, and mean in one line, plus a count of missing values if there are any. Run on the whole data frame instead of one column, it repeats this for every column at once, which is a fast way to eyeball the whole dataset in one go.

Look at `gender` and `smoking` in the output above. `gender` is a plain number, 1 or 2, standing in for a category. `smoking` is already text, 'Never', 'Previous', or 'Current'.

### When R guesses the wrong type

`read_csv()` decides each column's type by looking at the values in it. Most of the time that guess is right. It goes wrong in the same situation the messy spreadsheet exercise built on purpose: a column that's plain numbers for every patient except one, who was recorded as `'195 lbs'` instead of a number in kilograms. R has no way to know that `'lbs'` should be split off, converted, and the rest treated as a number. It sees text and stores the whole column as text. This is why the clean version of that spreadsheet needed every patient converted to the same unit, with that unit moved into the column *name* (`weight_kg`) and out of every individual *value*: it's the only way `read_csv()` ends up with a column it can calculate on.

### Spotting missing data

`is.na()` checks each value in a vector and returns `TRUE` where it's missing, `FALSE` where it isn't. `sum()` on that result counts the `TRUE`s, since R treats `TRUE` as 1 and `FALSE` as 0 when we add them up.

```{r}
#| label: missing-vpa-w
sum(is.na(nhanes$vpa_w))
```

That's how many participants have no value recorded for `vpa_w`, vigorous physical activity at work, out of `r format(nrow(nhanes), big.mark = ',')` participants in total. `nrow()`, from base R, returns the row count on its own, the same number `dim()` gives as its first value above, useful whenever we want the total without the column count. Whatever we go on to calculate from `vpa_w` only ever describes the participants who have a value. We should say so out loud whenever we report it, instead of letting it pass silently.

### Using the dictionary file

`nhanes_mph` comes with a dictionary which names every column and states what it means. Reading it prints the same column specification message we just unpacked, saying two columns of text this time instead of twenty-seven columns of mixed types. This book hides it from here on, since it says the same kind of thing every time and we've now read one properly. You'll still see it in your own console, and it's still worth a glance:

```{r}
#| label: read-dictionary
#| message: false
dictionary <- read_csv(here('data', 'mph_nhanes_20241107_dictionary.csv'))
dictionary
```

The dictionary's first column is called `Variable names`, with a space in it, so referring to it needs backticks, as Chapter 1 covered. `%in%`, from the same chapter, tests whether each value is a member of a set. Put inside square brackets after a data frame's column, with the row's index implied, this keeps only the rows where the condition is true:

```{r}
#| label: dictionary-lookup
dictionary[dictionary$`Variable names` %in% c('age', 'chol', 'smoking'), ]
```

That's the habit to build now: when a column's meaning isn't obvious from its name, the answer lives in this file, not in a guess. Chapter 3 uses this same file to build proper factors out of six of these columns.

## A first analysis, and your first script

### One summary statistic

```{r}
#| label: summary-chol
mean(nhanes$chol, na.rm = TRUE)
```

`chol` is total cholesterol, in mmol/L, and `r format(sum(is.na(nhanes$chol)), big.mark = ',')` participants have no value recorded for it. `na.rm = TRUE` drops those missing values before averaging the rest, as it did back in Chapter 1's `mean()` example on our own clinic log. Leave it out and the whole calculation returns `NA`, since R won't silently guess what a missing value should have been.

### A first taste of a grouped summary

`nhanes$chol` pulls the whole cholesterol column out as a bare vector, as `$` did in Chapter 1. Adding square brackets straight after it, with a condition inside, keeps only the values where that condition is `TRUE`:

```{r}
#| label: grouped-summary-preview
summary(nhanes$chol[nhanes$smoking == 'Never'])
summary(nhanes$chol[nhanes$smoking == 'Current'])
```

That's a grouped summary, built from tools we already have: `$`, `==`, and now `[` and `]` with a condition inside it. It works, but it's clunky already with just two groups, and `smoking` has three. Chapter 3 introduces `group_by()` and `summarise()`, a far nicer way to do this for as many groups as we like, in one line instead of several.

### A first taste of a plot

```{r}
#| label: plot-preview
#| message: false
library(ggplot2)

ggplot(nhanes, aes(x = chol)) +
  geom_histogram(binwidth = 0.5)
```

`ggplot2` code is joined with `+`, not the pipe. That's a real inconsistency in R, not a mistake in this book, and Chapter 3 explains why and unpacks the rest of the grammar properly. For today, just notice the shape: cholesterol piles up around 4 to 6 mmol/L, with a long tail of higher values trailing off to the right.

### That warning above the plot

The plot came with a line starting `Warning:`, saying it removed `r format(sum(is.na(nhanes$chol)), big.mark = ',')` rows. Read it, because we'll see this one a lot: this is what it says, and when it matters.

Firstly, a warning is not an error. An error means R gave up and produced nothing. A warning means R did the job and is telling us something about how. The plot is there and it's fine.

Secondly, look at the number. It's `r format(sum(is.na(nhanes$chol)), big.mark = ',')`, the same count of participants with no cholesterol value that we found a moment ago. A histogram puts each participant somewhere along the x-axis, and a missing value has nowhere to go, so `ggplot2` left those participants out and told us how many it dropped.

**Why we leave these warnings switched on.** R can be told to hide them, and plenty of people do, because they look untidy. This book doesn't, and neither should we for now. That warning is a free count of how much of our data a plot is based on. Hide it, and a plot built on half our dataset looks the same as a plot built on all of it. It's the same reason we write `na.rm = TRUE` deliberately and say what it dropped, instead of typing it out of habit until it stops registering.

**When it does need attention.** The number is the thing to check, not the warning itself. Compare it against what we expect, the way we just did with `is.na()`. If it matches, carry on. If it's different than expected, stop and find out why before trusting anything the plot shows: usually it means the wrong column, an import that didn't go as planned, or an earlier step that threw away more rows than intended. And whenever we report a result, we should say how many observations it's based on and how many were missing. A warning R printed for us is not the same as a reader being told.

### From console to script

Everything we've done so far in this chapter has been one line at a time. That's the right way to *try* something: type it in the console, look at what comes back, fix it if it's wrong. But Chapter 1 already made the argument against leaving it there. We will need to do this again, and we will not remember what we typed.

So the working habit for the rest of this course is two steps, not one. Try the line in the console until it does what we want. Then move the line that worked into a script, and save it. The console is where we experiment; the script is what we keep.

To build one now:

1.  In RStudio, go to **File → New File → R Script**, or key in **Ctrl+Shift+N** or **Cmd+Shift+N**. A blank file opens in the source pane, top-left.
2.  Type the lines you want to keep. For a first script, that's the `library()` calls, the `read_csv()` line, one summary, and one plot.
3.  Save it into the `scripts/` folder you built in Chapter 1, with a name that follows its conventions: no spaces, no `&` or `#`, something that says what it does. `first_analysis.R` is fine. `Untitled1.R` is not.

A complete first script looks like this, start to finish. Nothing in it is new; it's the pieces from earlier in this chapter, collected in the order they need to run:

``` r
library(readr)
library(here)
library(ggplot2)

nhanes <- read_csv(here('data', 'mph_nhanes_20241107.csv'))

mean(nhanes$chol, na.rm = TRUE)
sum(is.na(nhanes$chol))

ggplot(nhanes, aes(x = chol)) +
  geom_histogram(binwidth = 0.5)
```

Notice the shape of it. The `library()` calls come first, because nothing below them works until the packages are loaded. The `read_csv()` line comes next, because every line after it needs `nhanes` to exist.

### Proving it works

A script that runs on your machine right now, in a session where you've been typing for an hour, is not yet proof of anything. Half of what it needs might already be sitting in memory from something you typed and forgot.

So test it properly. In RStudio, go to **Session → Restart R**. That throws away every object and every loaded package, leaving you with an empty session, like the one a colleague, a marker, or you-in-six-months would start from. Now run your script from the top.

If it runs clean, the script is self-contained. If it fails, the error tells us precisely what we'd forgotten to write down, and the fix belongs *in the script*, not typed into the console again.

Chapter 3 starts from this point, with `nhanes` already read in and ready.

## Exercises

**Exercise 1** (ILO 2). This `read_csv()` call is supposed to read the dictionary file but doesn't work:

``` r
read_csv(here('data', 'mph_nhanes_dictionary.csv'))
```

Work out what's wrong with the path and write the corrected call.

**Exercise 2** (ILO 3). Check `mpa_w` (moderate physical activity at work): find its type with `class()`, and count how many values are missing with `is.na()` and `sum()`.

**Exercise 3** (ILO 4). Produce one summary statistic and one plot for `wal_r` (walking, hours/day, outside work). Remember `na.rm = TRUE` if it turns out to have missing values.

**Exercise 4** (ILO 4). Collect your answer to Exercise 3 into a proper script. Open a new R script, and write into it everything needed to go from a completely empty session to that summary and that plot: the `library()` calls, the `read_csv()` line, then the summary and the plot themselves. Save it into your `scripts/` folder with a sensible name.

Then test it the way this chapter described. Go to **Session → Restart R**, and run your script from the top. If it errors, read the message, work out what the fresh session was missing, and add that line to the script instead of typing it into the console. Repeat until it runs clean from a restart. That is the point of the exercise, not the summary itself.

## Self-check

Five questions on the concepts this chapter leans on hardest: paths, file types, and reading a real import problem. Answers are in the Solutions section below.

**1.** `read_csv(here('data', 'mhp_nhanes_20241107.csv'))` fails, even though `mph_nhanes_20241107.csv` really is in the `data` folder. Which of the four import failures is this?

a)  The file is the wrong type
b)  A path pasted with backslashes
c)  A wrong filename, here a typo (`mhp` instead of `mph`)
d)  The file is in the wrong folder

**2.** What's the actual difference between `dim(nhanes)` and `names(nhanes)`?

a)  They return exactly the same information in two formats
b)  `dim()` gives the number of rows and columns; `names()` lists every column name
c)  `names()` gives row and column counts; `dim()` lists column names
d)  `dim()` only works after `names()` has been run first

**3.** `sum(is.na(nhanes$vpa_w))` returns `7354`. What does that number mean?

a)  The sum of all values of `vpa_w`
b)  7,354 participants have `vpa_w` recorded as exactly zero
c)  7,354 participants have no value recorded for `vpa_w`
d)  There are 7,354 participants in the dataset in total

**4.** `read_csv()` is used to read both `mph_nhanes_20241107.csv` and `mph_nhanes_20241107_dictionary.csv`, in two separate calls. Why two calls, and not one?

a)  `read_csv()` can only read one row at a time
b)  They're two separate files, each holding a different table, so each needs its own call
c)  The dictionary file is not really needed, but the call is included anyway
d)  `read_csv()` automatically finds and reads both once you name one of them

**5.** Reading `05-prep2-messy-clinic.xlsx` with `read_csv()` produces a data frame with 1 column and 0 rows, and no error message. What does that tell you?

a)  The file is empty
b)  `read_csv()` refuses to read `.xlsx` files and stops immediately
c)  `.xlsx` is the wrong file type for `read_csv()`, and the lack of an error is why you always check what came in
d)  The import worked correctly and the file genuinely has one column

## Writing exercise

Using the mean and the histogram from Exercise 3, write a short paragraph, about 60 words, describing `wal_r` as you would in the results section of a report. State the mean and its units, say how many participants have no value recorded, and describe the shape of the distribution the histogram shows. This is the smallest version of a habit the rest of this book builds on: a number on its own, with no sentence around it, is not yet a result.

::: {.callout-note collapse='true'}
## Solutions

**Spreadsheet exercise**

The one problem not named in the list above is colour used to carry information: a block of rows is filled in yellow to mean 'smoker', with no `Smoker` column recording it at all, and the only place that says what the colour means is a caption sitting several rows away, disconnected from the table itself. R reads cell values, not cell fills, so the colour disappears completely on import, and the caption comes in as an isolated row of text with nothing left to connect it back to the patients it was describing.

The fixes needed, matching every problem in the sheet:

- Split the merged title row and the merged 'Measurements' sub-header back into one header row naming every column individually.
- Convert every patient's weight and height to the same units, kilograms and metres. Then move those units out of the values and into the column names instead (`weight_kg`, `height_m`), leaving plain numbers behind.
- Add a proper `Smoker` column holding an actual word ('Yes'/'No'), decoded from which rows were highlighted and what the caption said the highlighting meant.
- Delete both blank spacer rows.
- Delete or relocate the stray note about Patient 5's missed follow-up. It belongs in its own column, or nowhere in this table at all.
- Either delete the second table (the follow-up schedule) or save it as its own, separate CSV file. One sheet should hold one table.

**Exercise 1**

```{r}
#| label: sol-ex1
#| message: false
read_csv(here('data', 'mph_nhanes_20241107_dictionary.csv'))
```

The folder and filename were both fine. The problem was the letters `mph` typed as `mhp`.

**Exercise 2**

```{r}
#| label: sol-ex2
class(nhanes$mpa_w)
sum(is.na(nhanes$mpa_w))
```

**Exercise 3**

```{r}
#| label: sol-ex3
#| message: false
mean(nhanes$wal_r, na.rm = TRUE)
sum(is.na(nhanes$wal_r))

library(ggplot2)
ggplot(nhanes, aes(x = wal_r)) +
  geom_histogram(binwidth = 0.5)
```

**Exercise 4**

The finished script, in full. Saved as `scripts/wal_r_analysis.R`:

``` r
library(readr)
library(here)
library(ggplot2)

nhanes <- read_csv(here('data', 'mph_nhanes_20241107.csv'))

mean(nhanes$wal_r, na.rm = TRUE)
sum(is.na(nhanes$wal_r))

ggplot(nhanes, aes(x = wal_r)) +
  geom_histogram(binwidth = 0.5)
```

The two lines people most often leave out are the ones that were already true before they started: `library(readr)` and the `read_csv()` call. Both had been run earlier in the session, so `nhanes` existed and `read_csv()` worked. Neither felt like part of 'the analysis.' After a restart, neither is true any more, and the script fails on its first real line with `could not find function 'read_csv'` or `object 'nhanes' not found`, both of which Chapter 1 covered. That's what the restart is for: it finds the assumptions we didn't know we were making.

**Self-check answers**

1.  **c.** `mhp` versus `mph` is a plain typo in the filename, not a folder problem, a file-type problem, or a backslash problem.
2.  **b.** `dim()` returns row and column counts. `names()` returns the column names themselves. Neither substitutes for the other.
3.  **c.** `is.na()` marks each missing value `TRUE`, and `sum()` adds those up. The number is a count of participants with no recorded value, nothing about the values that are present.
4.  **b.** `mph_nhanes_20241107.csv` and `mph_nhanes_20241107_dictionary.csv` are two separate files on disk, each holding a different table, so each needs its own `read_csv()` call.
5.  **c.** `.xlsx` is a zip archive of XML files, not a CSV, so `read_csv()` produces nonsense, not a real dataset. The absence of an error is the trap: always check what came in after reading a file, not just whether it 'worked.' `dim()` and `glimpse()`, both used earlier in this chapter, are two ways to do that.

**Writing exercise (sample answer)**

Participants reported a mean of 0.21 hours/day of walking outside work (`wal_r`), based on the 34,418 participants with a value recorded, with 7,347 participants who had a missing value. The distribution has a long tail to the right: most participants reported no walking outside work at all, and a small number reported several hours a day, pulling the mean above what a typical participant's value looks like.
:::