# Describing and visualising data (Week 7) {#sec-07-descriptive}

Scrolling through a dataset of tens of thousands of participants tells us almost nothing. Twenty rows we can read; twenty thousand we can't hold in our head at all. Chapter 2 got a dataset of that size into R and confirmed every column had arrived, which isn't the same as knowing what any of it looks like. Seeing that takes summary statistics and graphs, and building them is the work of this chapter.

Which statistic is the right one depends entirely on the kind of variable in front of us. A mean and a standard deviation (SD) describe a continuous variable and say nothing sensible about a categorical one, where a count and a percentage do the job instead. So the types come first here, and the summaries, the plots, and the choice of what to show all follow from getting them right. The chapter's destination is a 'Table 1', the participant-characteristics table that opens most published health research, along with the paragraph of prose that has to go with it.

## Chapter intended learning outcomes

By the end of this week you will be able to:

1. Classify a variable as nominal, ordinal, binary, count, or continuous, and explain how variable type constrains the analysis available, converting a coded variable to a factor with `factor()` and setting its reference level with `relevel()`
2. Select and construct an appropriate table or plot for one or more variables of a given type, using `table()`, `prop.table()`, `ggplot2`, and `gtsummary::tbl_summary()`
3. Calculate and interpret measures of central tendency and spread, and justify a choice between mean/SD and median/IQR by distributional shape, preparing the data with `dplyr`'s `select()`, `filter()`, `mutate()`, and `arrange()` and summarising it with `group_by()` and `summarise()`
4. Describe the normal distribution and apply the 68-95-99.7 rule
5. Critique a published figure and identify design choices that mislead

## The data: `nhanes_mph`

NHANES is the National Health and Nutrition Examination Survey, run by the National Center for Health Statistics, part of the US Centers for Disease Control and Prevention. It's designed to describe the health of the US population, and it's unusual in that it doesn't just ask people questions. Participants are interviewed at home and then examined in a mobile clinic, where measurements are taken and blood samples drawn. That's why our data can carry a measured cholesterol value, not only a self-report answer.

`nhanes_mph` is a curated extract of it, prepared specifically for this course: 41,765 adults aged 18 to 85, described by 27 variables. Those cover demographics, education, smoking and alcohol, physical activity split by domain, sedentary time, total cholesterol, depression symptoms, and whether the participant died during follow-up. The full list is in the dictionary file, which we read in alongside the data below and use throughout.

### One simplification

NHANES is not a simple random sample of Americans. It's a **complex survey**. Some groups are deliberately sampled more heavily than others in certain years, and participants are recruited in geographic clusters, not one at a time. Every participant also carries a survey weight, saying how many people in the population they stand for. Ideally, all of that would go into the calculation, to get an estimate representative of the US population. This course treats `nhanes_mph` as though it were a simple cohort instead: every participant counts once, and counts the same. That's a real simplification, and it means the numbers we produce here would not be the right numbers to publish as population estimates. Survey weights will not be covered anywhere in the MPH curriculum, not deferred to Advanced Statistics.

## Variable types

Every variable falls into one of two broad kinds, categorical and numeric.

A **categorical** variable puts each participant into one of a fixed set of groups. `gender` and `eth` are **nominal**: the categories have no natural order. `phq.c`, depression classification, is **ordinal**: normal, mild symptoms, and possible depression run in a genuine order, even though the gap between them isn't a measured distance. `blm.cvd`, baseline cardiovascular disease, is **binary**: a two-category special case of categorical, present or absent, even though in the data it takes the value 0 or 1.

A **numeric** variable is a number, and numbers come in two shapes that matter differently for analysis. `phq`, the raw depression symptom score behind `phq.c`, is a **count** (sometimes called discrete): whole numbers only, 0 through 6. `chol`, `age`, and `sed` are **continuous**: any value in a range is possible, 4.7 mmol/L as easily as 4.71.

The variable type decides the summary and the plot. A mean and an SD, or a median and an interquartile range (IQR), describe a **continuous** variable. All four are defined properly in the Grouped summaries section below. A **categorical** variable, however few its categories, is summarised as a proportion or a frequency table instead: for a binary variable like `blm.cvd`, the proportion of 1s and the mean of its 0/1 values come out as the same number, but we still call it a proportion, since what it summarises is a category, not a measurement. This book says *continuous* from here on when that's specifically what's meant, not the vaguer *numeric*, since numeric alone doesn't say whether a variable is a count or continuous.

## Setup

As in Chapter 2, we read the data with `read_csv()` from the `readr` package. We build the file path with `here()` from the `here` package. This builds a path from the project folder down to the file, so the code runs on any machine no matter where the project sits. `nhanes_mph` arrives as two files: the data, and a separate dictionary that describes every variable. Both are plain CSV files, so we read both with `read_csv()`.

This chapter loads one package new to the book: `gtsummary`, for building tables. `dplyr`, for manipulating data, was loaded back in Chapter 1, and `ggplot2`, for plotting, in Chapter 2, so both are already familiar. Installing and loading it are still the two separate actions Chapter 1 distinguished, so if `gtsummary` isn't installed already, run this once in the console. You don't need to put it in the script.

``` r
install.packages('gtsummary')
```

After that, the `library()` calls below are all we need, this session and every session after it.

```{r}
#| label: setup
#| message: false

library(readr)
library(dplyr)
library(ggplot2)
library(gtsummary)
library(here)

nhanes <- read_csv(here('data', 'mph_nhanes_20241107.csv'))
dictionary <- read_csv(here('data', 'mph_nhanes_20241107_dictionary.csv'))
```

```{r}
#| label: setup-hidden
#| include: false
options(width = 76)
source(here('R', 'gtsummary-knitprint.R'))
```

Running that in your own console prints more than it does on this page. The `library()` calls announce themselves, and each `read_csv()` prints the column specification message Chapter 2 unpacked. This book hides both from here on, as both have been explained.

`glimpse()`, from `dplyr`, met in Chapter 2, shows every column, its type, and its first few values. It's useful for checking what came in, not what we expected to come in.

```{r}
#| label: glimpse-raw
glimpse(nhanes)
```

Six columns describe a category but aren't usable as one yet: `gender`, `eth`, `edu`, `smoking`, `phq.c`, and `blm.cvd`. `glimpse()` reports how R has **stored** each column, which is a different question from what the variable statistically **is**. `blm.cvd` is binary, so categorical, but R read its 0s and 1s as numbers and lists it as `<dbl>`; `gender` and `edu` are the same, plain integers standing in for a label. `smoking` and `phq.c` are categorical too, just stored as text: `glimpse()` lists both as `<chr>`, not numbers at all. Whichever way R happened to store a column, the variable's statistical type doesn't change, and R can't infer one from the other just by looking at it. Whatever the raw form, we should consult the data documentation, e.g. the data dictionary, whenever possible.

```{r}
#| label: dictionary-lookup
dictionary |>
  filter(`Variable names` %in% c('gender', 'eth', 'edu', 'smoking', 'phq.c', 'blm.cvd'))
```

The dictionary gives a numeric code against each category for `gender`, `eth`, `edu`, and `blm.cvd`: `1=Male; 2=Female`, and so on. Those are the four we have to translate. For `smoking` and `phq.c` it lists the categories, and `glimpse()` already told us why no translation is needed: they don't arrive as bare numbers at all. `smoking` is already the text 'Never', 'Previous', or 'Current'. `phq.c` is already a label too, with the original code number kept at the front of the string, for example `'0 Normal'`. The dictionary isn't exhaustive, either. `dth`, the death indicator we use in Chapter 7, has not had a coding given to it. But by programming convention, `0` is no (or the logical `FALSE`), and `1` is yes (or the logical `TRUE`).

We turn every one of the six into a factor with `mutate()` and `factor()`, whether the raw values are numbers or text. `factor()` takes the raw values (`levels`) and the labels we want to give them (`labels`), in matching order. The labels are ours to choose, and we take one small liberty below: the dictionary abbreviates `eth`'s first category to 'Mexican', and we write it out as 'Mexican American', the full NHANES category name. One of the six, `phq.c`, is ordinal: its categories run from normal to possible depression in a natural order, so we build it with `ordered = TRUE`. `gender`, `eth`, `smoking`, and `blm.cvd` are nominal, with no natural order, so they're plain factors for the more familiar reason. `edu` is the odd one out. Its three levels do run in a natural order, but we build it as a plain factor anyway, because the regression models in Chapters 6 and 7 compare each education level with a reference level, one at a time, and don't use the order at all. By default, the first level listed becomes the *reference* level, the category everything else gets compared against. The next section says more about what that means and why it's set deliberately.

```{r}
#| label: recode

nhanes <- nhanes |>
  mutate(
    gender = factor(gender,
      levels = c(1, 2),
      labels = c('Male', 'Female')),
    eth = factor(eth,
      levels = c(1, 2, 3, 4, 5),
      labels = c('Mexican American', 'Other Hispanic', 'White', 'Black', 'Others')),
    edu = factor(edu,
      levels = c(1, 2, 3),
      labels = c('Less than high school', 'High school graduate', 'Above high school')),
    smoking = factor(smoking,
      levels = c('Never', 'Previous', 'Current')),
    phq.c = factor(phq.c,
      levels = c('0 Normal', '1 Mild symptoms', '2 Possible depression'),
      labels = c('Normal', 'Mild symptoms', 'Possible depression'),
      ordered = TRUE),
    blm.cvd = factor(blm.cvd,
      levels = c(0, 1),
      labels = c('No', 'Yes'))
  )

glimpse(nhanes)
```

The blank and missing values in `smoking`, `edu`, `phq.c`, and `blm.cvd` have become `NA` in the process, as they were before recoding. `factor()` relabels the values that are present and carries missingness through unchanged. From here on, every chapter opens with a short setup chunk that repeats these same steps, so each chapter stands on its own.

## Checking and setting a reference level

The reference level decides which category a table or a chart reads everything else against. It's a reporting choice before it's a modelling one, so we set it deliberately from here, not only once a categorical variable enters a regression model in Chapter 6. Let's examine the levels in a factor. `levels()`, from base R, lists a factor's categories in their current order:

```{r}
#| label: levels-smoking
levels(nhanes$smoking)
```

`'Never'` is first, so `'Never'` is the reference. Everything else gets read against it: 'Previous' means previous compared with never, and 'Current' means current compared with never. That's the sensible choice here, and it's what we asked for when we wrote `levels = c('Never', 'Previous', 'Current')` in the recode above. Change the order in that line and we change the reference.

Sometimes we want a different reference without rebuilding the factor. `relevel()`, also from base R, takes a factor and moves the level named in `ref` to the front:

```{r}
#| label: relevel-eth
eth_white_ref <- relevel(nhanes$eth, ref = 'White')
levels(eth_white_ref)
```

`'White'` has moved to the front, so it would now be the reference, with the other four groups read against it. The rest keep their order, shifted along by one.

Notice we assigned the result to a new object, `eth_white_ref`, instead of writing it back into `nhanes`. Every chapter in this book rebuilds `nhanes` with the same recode. If we changed this factor here, this chapter's `nhanes` would stop matching the one later chapters build, and a table or model would come out different for no visible reason. When we're demonstrating something, we work on a copy.

Right now the reference level only decides the order categories appear in a table, or along a bar chart's axis. Once a categorical variable enters a regression model in Chapter 6, it decides what every coefficient means.

## Manipulating a dataset: the `dplyr` verbs

A public health question is almost never about everyone in a dataset at once. Four verbs from `dplyr` cover most of what we need to get from the raw dataset down to the subgroup, the columns, and the order a question asks about.

`select()` keeps only the named columns:

```{r}
#| label: select-demo
nhanes |> select(age, gender, sed, chol)
```

`filter()` keeps only the rows that satisfy a condition. Here, that's participants under 40:

```{r}
#| label: filter-demo
nhanes |> filter(age < 40)
```

`mutate()` adds a new column, or changes an existing one, as we already saw above in the recoding step. Here it adds a new column that classifies participants as under or over 65:

```{r}
#| label: mutate-demo
nhanes |> mutate(over_65 = age >= 65) |> select(age, over_65)
```

`arrange()` sorts the rows by a column, lowest to highest by default:

```{r}
#| label: arrange-demo
nhanes |> arrange(age)
```

These four combine in a short pipe to answer a real question. For example, who are the ten oldest current smokers in the sample?

```{r}
#| label: verbs-combined
nhanes |>
  filter(smoking == 'Current') |>
  arrange(desc(age)) |>
  select(age, gender, smoking) |>
  slice(1:10)
```

`desc()` reverses the sort order. `slice()`, also from `dplyr`, keeps only the rows at the given positions. Here, that's the first ten after sorting.

## Grouped summaries

`group_by()` splits a dataset into groups without changing how it looks. `summarise()` then collapses each group down to one row of statistics. Used together, they answer questions like this: what's average cholesterol, separately for men and women?

```{r}
#| label: group-summarise
nhanes |>
  filter(!is.na(gender), !is.na(chol)) |>
  group_by(gender) |>
  summarise(
    mean_chol = mean(chol, na.rm = TRUE),
    sd_chol = sd(chol, na.rm = TRUE),
    median_chol = median(chol, na.rm = TRUE),
    q1_chol = quantile(chol, probs = 0.25, na.rm = TRUE),
    q3_chol = quantile(chol, probs = 0.75, na.rm = TRUE),
    n = n()
  )
```

`filter()` here takes two conditions separated by a comma, which means the same as `&`: a row only stays if both are true. That's a shorter way to write `filter(!is.na(gender) & !is.na(chol))`, and it's the form used throughout this book whenever several conditions apply at once.

Four of those functions are new. `mean()` we already met in Chapter 1. `sd()`, from base R, returns the standard deviation: a measure of how spread out the values are, in the same units as the variable itself. `median()`, also base R, returns the middle value once everything is sorted, so half the participants sit below it and half above. `quantile()`, also base R, calculates any quantile we ask for: `probs = 0.25` and `probs = 0.75` ask for the first and third quartile, the two values the interquartile range sits between. There's a separate function, `IQR()`, that returns only the gap between them as a single number, in the same units as the variable itself, instead of the two quartiles. This book reports the two quartiles, not that gap: the gap on its own doesn't say *where* the middle half of the data sits, and two variables with the same gap can sit in completely different places.

`na.rm = TRUE` tells the functions to drop missing values before calculating. Without it, any `NA` in a column makes the whole result `NA`. The `filter()` step above drops rows where `gender` or `chol` itself is missing, so the group sizes reported in `n` reflect the rows the means are calculated from. `n()`, used inside `summarise()`, counts the rows in each group.

The `gender` half of that filter turns out to change nothing: every participant in this extract has a gender recorded. The `chol` half of the same filter matters: `r format(sum(is.na(nhanes$chol)), big.mark = ',')` participants have no cholesterol value, and without it they'd be counted in `n` while contributing nothing to the mean.

Whether we report the mean and SD or the median and IQR depends on shape. A **skewed** distribution, one whose tail stretches much further on one side than the other, gets misrepresented by a mean, because the mean gets pulled toward the tail. A rule of thumb is to look at the shape before choosing. We'll build the plot we need for that in a moment, but categorical variables need a different kind of summary first: a count, not a mean.

## Frequency and cross-tabulations

`table()`, from base R, counts how many rows fall into each category of a variable. Given one variable, it produces a **frequency table**:

```{r}
#| label: table-smoking
table(nhanes$smoking)
```

`table()` drops `NA` values by default, not counting them as their own category, so this count is out of everyone with a recorded smoking status, not out of everyone in the dataset. Raw counts are hard to compare across groups of different sizes, so we usually want percentages instead. `prop.table()`, also from base R, takes a table and turns its counts into proportions of the total:

```{r}
#| label: proptable-smoking
table(nhanes$smoking) |> prop.table()
```

Multiplying by 100 and rounding turns those proportions into percentages, in whatever way is easiest to read.

Giving `table()` two variables instead of one produces a **cross-tabulation**: a two-way table with one variable's categories as rows and the other's as columns, one cell per combination. Here's smoking status by gender:

```{r}
#| label: table-smoking-gender
smoking_gender <- table(nhanes$smoking, nhanes$gender)
smoking_gender
```

The rows are `smoking`, the columns are `gender`, because that's the order the two variables were given to `table()` in. As with the frequency table above, raw counts are easier to compare as percentages, but now there's a choice to make: percentage of what? `prop.table()` takes a `margin` argument for this. `margin = 1` divides each row by its own row total, giving **row percentages**, the breakdown of gender within each smoking category:

```{r}
#| label: proptable-row
prop.table(smoking_gender, margin = 1)
```

`margin = 2` divides each column by its own column total instead, giving **column percentages**, the breakdown of smoking status within each gender:

```{r}
#| label: proptable-col
prop.table(smoking_gender, margin = 2)
```

Both tables are built from the same counts, but they answer different questions, and the research question decides which one belongs in a report. This is the same two-by-two logic the epidemiology half already uses for exposure-by-outcome tables, just with more than two categories on one side.

## Plotting with `ggplot2`

Every `ggplot2` plot is built from the same three pieces: the **data**, the **aesthetic mapping** (`aes()`, which decides which variables go on which axis or colour), and the **geometry** (`geom_*()`, which decides what kind of shape represents the data). These pieces are joined with `+`, not the pipe `|>` we've been using so far, because `+` does a different job. `x() |> y()` passes the output of `x()` as an input to `y()`. `+` in `ggplot2` means **layering** instead, an idea introduced by Leland Wilkinson's book *The Grammar of Graphics*: a plot is built as layers stacked on top of one another, data first (`ggplot(nhanes, ...)` below), then the aesthetic mapping (`aes(x = sed)` below), then the geometry (`geom_histogram()` below).

```{r}
#| label: histogram-demo
ggplot(nhanes, aes(x = sed)) +
  geom_histogram(binwidth = 1) +
  labs(x = 'Sedentary time (hours/day)', y = 'Count')
```

`ggplot()` names the data and the mapping, `geom_histogram()` says to draw it as a histogram, and `labs()` sets the axis labels. Use `labs()` every time: left out, the axes are labelled with the raw column name, so this one would have read `sed` instead of 'Sedentary time (hours/day)', and anyone outside the course would have to guess what that means.

That warning about removed rows is the one Chapter 2 explained: `sed` has `r format(sum(is.na(nhanes$sed)), big.mark = ',')` missing values, so `ggplot2` left them out and said so. It will appear above most plots in this chapter, with a different number each time depending on the variable. Glance at the number, check it's the size we expect, and carry on.

`ggplot2` is highly versatile and can produce nearly any plot we need. Some examples below:

A bar chart of smoking status:

```{r}
#| label: barchart-demo
ggplot(nhanes |> filter(!is.na(smoking)), aes(x = smoking)) +
  geom_bar() +
  labs(x = 'Smoking status', y = 'Count')
```

`geom_bar()` is like a histogram for a categorical variable, another way to represent the frequency table we built above. The code is very similar.

A boxplot of cholesterol by gender:

```{r}
#| label: boxplot-demo
ggplot(nhanes |> filter(!is.na(gender)), aes(x = gender, y = chol)) +
  geom_boxplot() +
  labs(x = 'Gender', y = 'Total cholesterol (mmol/L)')
```

This is useful for examining the relationship between a continuous variable and a categorical one. The code is similar, but with two differences: it explicitly filters out participants with missing gender, and it names both the x and y axis variables. `geom_boxplot()` replaces the `geom_histogram()` from above.

A scatterplot of cholesterol against age:

```{r}
#| label: scatter-demo
ggplot(nhanes, aes(x = age, y = chol)) +
  geom_point(alpha = 0.1) +
  labs(x = 'Age (years)', y = 'Total cholesterol (mmol/L)')
```

`geom_point()` is useful to examine the relationship between two continuous variables. `alpha = 0.1` makes each point mostly transparent. Where thousands of points overlap, the darker regions show where the data are dense. A plain scatterplot of this many points would otherwise be a solid block of ink.

::: {.callout-note}
## Try it yourself: bin width and scale

Take the histogram of `sed` above. Change `binwidth` to `0.1`, then to `5`, and re-run it each time. Then add `+ ylim(0, 500)` to the original plot. `ylim()`, from `ggplot2`, fixes the lower and upper limits of the y-axis by hand instead of letting the plot choose them to fit the data. Be careful with it: anything falling outside the limits you set is dropped from the plot altogether, not drawn cut off at the edge, so a bar taller than the limit doesn't get shortened, it disappears. For each change, write one sentence on what altered in the plot, and one sentence on what the data themselves did not change. This is the same distinction the interactive lecture drew between a plot design choice and the underlying distribution.

This is the first of several 'Try it yourself' boxes in the book, and none of them has a written solution. Running the code is the answer.
:::

## A first Table 1

A 'Table 1' is a common first table in a health research report. It shows participant characteristics, often split by a grouping variable. We could do it manually with the descriptive statistics commands above (e.g. `mean()`, `table()`) and it is a good exercise. However, the more convenient way is to use `tbl_summary()`, from `gtsummary`.

The code below creates the summary table overall for the five selected variables.

```{r}
#| label: table1-demo
nhanes |>
  select(age, gender, eth, smoking, chol) |>
  tbl_summary()
```

Adding `by =` splits the table by a grouping variable, giving one column per group:

```{r}
#| label: table1-by-gender
nhanes |>
  select(age, gender, eth, smoking, chol) |>
  tbl_summary(by = gender)
```

`tbl_summary()`'s default display for a continuous variable is the median with the first and third quartiles, which the table footnotes as `Median (Q1, Q3)`. To change it to mean and SD, we tell it what statistic to use with the `statistic` argument:

```{r}
#| label: table1-by-gender-2
nhanes |>
  select(age, gender, eth, smoking, chol) |>
  tbl_summary(
    by = gender,
    statistic = list(all_continuous() ~ '{mean} ({sd})')
  )
```

Three things in that line are new. `list()`, from base R, bundles several settings together, one per variable or group of variables, which is what `statistic` needs when there's more than one display rule to give. `all_continuous()`, from `gtsummary`, means 'every continuous variable in this table', so one line covers both `age` and `chol` at once instead of naming each separately. The `~` between it and the text that follows pairs the two: which variables, and what to show for them. This book properly introduces `~` as formula syntax in Chapter 4; here it's doing a related but different job, pairing a selection with a display rule instead of pairing the variable being modelled with the one it's modelled on. And `'{mean} ({sd})'` is a piece of text with slots in it. `gtsummary` works out the mean and the standard deviation for each variable, then drops each number into the spot where its name appears in curly braces. It comes out as something like `47.3 (18.1)`. Everything outside the braces, here a space and a pair of round brackets, is printed as written. `tbl_summary()` is highly customisable. We can change the statistic for any specific variable, or format different variables differently. That's outside the scope of this course, and left as an exploration exercise for anyone interested.

## Worked example: describing age

How old are the participants in this sample, and how much does that vary?

```{r}
#| label: worked-summary
nhanes |>
  summarise(
    mean_age = mean(age, na.rm = TRUE),
    sd_age = sd(age, na.rm = TRUE),
    median_age = median(age, na.rm = TRUE),
    q1_age = quantile(age, probs = 0.25, na.rm = TRUE),
    q3_age = quantile(age, probs = 0.75, na.rm = TRUE)
  )
```

```{r}
#| label: worked-histogram
ggplot(nhanes, aes(x = age)) +
  geom_histogram(binwidth = 5) +
  labs(x = 'Age (years)', y = 'Count')
```

The histogram is close to symmetric, with no strong tail in either direction. So the mean and SD are a fair summary here. Median and IQR would tell much the same story. Now let's split by gender:

```{r}
#| label: worked-boxplot
ggplot(nhanes |> filter(!is.na(gender)), aes(x = gender, y = age)) +
  geom_boxplot() +
  labs(x = 'Gender', y = 'Age (years)')
```

```{r}
#| label: worked-table1
nhanes |>
  select(age, gender) |>
  tbl_summary(by = gender)
```

Age looks similar across the two gender groups, both in the boxplot and in the Table 1. There's no obvious difference to report here.

## Exercises

**Exercise 1** (ILO 3). Using the four `dplyr` verbs, and a pipe, answer this: among participants aged 65 or over who have never smoked, what are the ten highest cholesterol values? Keep only `age`, `gender`, and `chol` in your answer. Then add a column called `chol_high` that is `TRUE` when `chol` is above 6 mmol/L and `FALSE` otherwise.

**Exercise 2** (ILOs 2, 3). Describe the distribution of sedentary time (`sed`). Report the mean and SD or the median and IQR, whichever the shape justifies, and support your choice with a histogram and a boxplot. Then produce a scatterplot of `chol` against `sed`, label both axes with their units, and write one sentence on whether the plot suggests any relationship between the two.

**Exercise 3** (ILOs 2, 5). Build a histogram of total cholesterol (`chol`) with a `binwidth` of 0.5, labelled properly with `labs()`. Then produce two more versions of it: one with a `binwidth` of 0.05, and one with the original bin width but with `+ ylim(0, 1000)` added. Write one sentence for each of the two changes saying what it does to the impression the plot gives, and one sentence saying which of the three you would put in a report and why.

**Exercise 4** (ILOs 1, 2, 3). Compare total cholesterol (`chol`) by gender. Produce a grouped summary and a boxplot, and write one sentence comparing the two groups. Then use `levels()` to state which gender is currently the reference level, and `relevel()` to make the other one the reference instead, assigning the result to a new object instead of overwriting `nhanes`.

**Exercise 5** (ILOs 1, 2). Produce a Table 1 of `age`, `gender`, `eth`, `smoking`, and `chol`, stratified by depression classification (`phq.c`).

**Exercise 6** (ILO 2). Build a cross-tabulation of smoking status (`smoking`) against education (`edu`) with `table()`. Produce both the row percentages and the column percentages with `prop.table()`.

Then answer this question using whichever of the two is appropriate: *among people who never smoked, how is education distributed?* State which of your two tables answers it, and say in one sentence why the other one doesn't. Finally, produce a bar chart of `edu`, labelled properly.

## Writing exercise

The course's assessment is a structured paper on a dataset you haven't seen, where every question asks for an analysis *and* an interpretation, so these exercises can be treated as formative assessments.

Using the Table 1 from Exercise 5, write a short paragraph of a maximum of 150 words. Describe how participant characteristics differ, or don't, across the three depression classification groups. State which characteristics look similar across groups and which look different, and avoid overstating a difference the table doesn't support.

Start by stating how many participants the table is based on, and how many were excluded because they had no depression classification recorded. The number is in the message printed above the table.

::: {.callout-note collapse='true'}
## Solutions

**Exercise 1**

```{r}
#| label: sol-ex1
nhanes |>
  filter(age >= 65, smoking == 'Never') |>
  arrange(desc(chol)) |>
  select(age, gender, chol) |>
  slice(1:10)
```

The comma inside `filter()` still means `&`, both conditions have to be true. Then `arrange(desc(chol))` sorts from highest cholesterol down, `select()` keeps the three columns asked for, and `slice(1:10)` takes the top ten rows.

The `mutate()` step is separate, because adding a column to ten hand-picked rows isn't much use. Applied to the whole dataset it is:

```{r}
#| label: sol-ex1-mutate
nhanes |>
  mutate(chol_high = chol > 6) |>
  select(age, chol, chol_high)
```

`chol > 6` is a comparison, like the ones in Chapter 1, and it returns `TRUE` or `FALSE` for every row. Participants with no cholesterol value recorded get `NA`, not `FALSE`, which is correct: we don't know whether they're above 6, and guessing 'no' would be inventing data.

**Exercise 2**

```{r}
#| label: sol-ex2
nhanes |>
  summarise(
    median_sed = median(sed, na.rm = TRUE),
    q1_sed = quantile(sed, probs = 0.25, na.rm = TRUE),
    q3_sed = quantile(sed, probs = 0.75, na.rm = TRUE)
  )

ggplot(nhanes, aes(x = sed)) +
  geom_histogram(binwidth = 1) +
  labs(x = 'Sedentary time (hours/day)', y = 'Count')

ggplot(nhanes, aes(x = sed, y = '')) +
  geom_boxplot() +
  labs(x = 'Sedentary time (hours/day)', y = NULL)
```

The histogram is right-skewed, with a long tail of participants reporting many hours of sedentary time. So the median and IQR describe it more fairly than the mean and SD would.

```{r}
#| label: sol-ex2-scatter
ggplot(nhanes, aes(x = sed, y = chol)) +
  geom_point(alpha = 0.1) +
  labs(x = 'Sedentary time (hours/day)',
       y = 'Total cholesterol (mmol/L)')
```

The scatterplot shows no clear relationship. The cloud of points is roughly the same height right across the range of sedentary time, not drifting up or down as we move left to right. Chapter 6 puts a number and a confidence interval on that impression instead of leaving it to the eye.

**Exercise 3**

```{r}
#| label: sol-ex3-binwidth
ggplot(nhanes, aes(x = chol)) +
  geom_histogram(binwidth = 0.5) +
  labs(x = 'Total cholesterol (mmol/L)', y = 'Count')

ggplot(nhanes, aes(x = chol)) +
  geom_histogram(binwidth = 0.05) +
  labs(x = 'Total cholesterol (mmol/L)', y = 'Count')

ggplot(nhanes, aes(x = chol)) +
  geom_histogram(binwidth = 0.5) +
  ylim(0, 1000) +
  labs(x = 'Total cholesterol (mmol/L)', y = 'Count')
```

Narrowing the bins to 0.05 breaks the smooth shape into a spiky, gappy outline. Cholesterol is recorded to a limited number of decimal places, so at this bin width many bins fall in the gaps between recorded values and come out empty. Those spikes are an artefact of how the values were written down, not a real feature of cholesterol.

Capping the y-axis at 1,000 does something worse. Every bar taller than 1,000 vanishes from the plot completely. The tallest bar here counts over 7,000 participants, and the whole middle of the distribution is taller than the cap, so what's left is the two tails with a hole where most of the data should be.

The first version is the one to put in a report. It shows the whole distribution at a bin width where the shape is readable, and it doesn't drop anything. Notice that all three plots are drawn from the same numbers: nothing about cholesterol changed between them, only the choices we made about how to draw it. That's the lecture's point about misleading figures.

**Exercise 4**

```{r}
#| label: sol-ex4-summary
nhanes |>
  filter(!is.na(gender)) |>
  group_by(gender) |>
  summarise(
    mean_chol = mean(chol, na.rm = TRUE),
    sd_chol = sd(chol, na.rm = TRUE),
    n = n()
  )

ggplot(nhanes |> filter(!is.na(gender)), aes(x = gender, y = chol)) +
  geom_boxplot() +
  labs(x = 'Gender', y = 'Total cholesterol (mmol/L)')
```

Mean cholesterol is slightly higher in women than in men, and the boxplots overlap substantially. That's a small difference in the centre, not a separation of the two distributions.

```{r}
#| label: sol-ex4-relevel
levels(nhanes$gender)

gender_female_ref <- relevel(nhanes$gender, ref = 'Female')
levels(gender_female_ref)
```

`'Male'` is listed first, so `'Male'` is currently the reference level, which follows from the recode listing the numeric code `1` first and labelling it `'Male'`. After `relevel()`, `'Female'` is at the front of the new object and would be the reference instead. `nhanes` itself is untouched.

**Exercise 5**

```{r}
#| label: sol-ex5
nhanes |>
  select(age, gender, eth, smoking, chol, phq.c) |>
  tbl_summary(by = phq.c)
```

Notice the message above the table: `4499 missing rows in the 'phq.c' column have been removed.` That's `gtsummary` telling us something important, not complaining. It isn't an error, and the table is fine.

What it means is this. We asked for the table split three ways, by `phq.c`. A participant with no depression classification recorded can't go in any of the three columns, and `tbl_summary()` left them out. The table describes the `r format(nrow(nhanes) - 4499, big.mark = ',')` who remain.

It matters here more than usual, because this is the table the writing exercise asks you to describe. Describing it as though it covered everybody would overstate what it shows by 4,499 people.

**Exercise 6**

```{r}
#| label: sol-ex6-table
smoking_edu <- table(nhanes$smoking, nhanes$edu)
smoking_edu

prop.table(smoking_edu, margin = 1)
prop.table(smoking_edu, margin = 2)
```

The row percentages are the ones that answer the question. `smoking` was given to `table()` first, so smoking is the rows, and `margin = 1` divides each row by its own total. The 'Never' row then reads as the education breakdown *within* people who never smoked, which is what was asked, and those three numbers add to 100%.

The column percentages answer a different question: within each education group, what share smoke, used to smoke, and never smoked. Both tables come from the same counts, so neither is more correct than the other in the abstract. What decides it is which total we want the percentages to be a share of, and that comes from the question, not from the data.

```{r}
#| label: sol-ex6-bar
ggplot(nhanes |> filter(!is.na(edu)), aes(x = edu)) +
  geom_bar() +
  labs(x = 'Education', y = 'Count')
```

**Writing exercise (sample answer)**

Depression classification was recorded for `r format(nrow(nhanes) - 4499, big.mark = ',')` of the `r format(nrow(nhanes), big.mark = ',')` participants, with the rest being excluded. Among those classified, age, gender, and ethnicity distributions were broadly similar across the three groups. Current smoking was more common among participants classified with possible depression than among those classified as normal. Median total cholesterol was similar across all three groups. These are descriptive comparisons only, based on group summaries and counts, not any formal test of difference. Hypothesis tests come in Chapter 5.
:::
