# Estimation, precision, and sample size (Week 8) {#sec-08-estimation}

Chapter 3 taught us to describe a variable with a mean and a standard deviation. Take a different sample from the same population and both numbers come out different. That wobble is measurable, and measuring it is what turns a number into an estimate we can defend. Sampling variability and the central limit theorem, from this week's interactive lecture, are what make it predictable enough to measure.

Two terms that sound alike carry the whole chapter, and they aren't interchangeable. The standard deviation describes the spread of the *data*. The standard error describes the precision of an *estimate*. Out of the second comes the confidence interval, or CI, which we calculate by hand before checking it against R's own functions, since a formula met once from the inside is harder to misread later. The same CIs cover the difference between two groups as well as a single mean, and each one has to be stated in the units of the problem: a CI nobody can put into a sentence hasn't been interpreted. The machinery also runs backwards. A CI narrows as the sample grows, so fixing the width we want tells us the sample we need, which is how a study gets sized before anyone has collected anything.

## Chapter intended learning outcomes

By the end of this week you will be able to:

1. Understand the difference between population and sample, and why statistical inference focuses on the population
2. Explain why a sample estimate varies from sample to sample, and state the practical consequence of the central limit theorem for the shape of that variation
3. Distinguish the standard deviation from the standard error, and calculate the standard error of a mean and of a proportion, using `sd()` and `sqrt()`
4. Construct a 95% confidence interval for a mean, a proportion, a difference in means, and a difference in proportions, by formula and with `t.test()` and `prop.test()`
5. Interpret a confidence interval correctly, and identify the common misinterpretation
6. Explain how sample size and variability jointly determine an estimate's precision, and calculate the sample size a target margin of error requires

## Setup

As in Chapter 3, we read the data with `read_csv()` and turn the six coded variables into factors against the dictionary. This chapter also brings in a second dataset for the first time: `strep_tb`. Unlike `nhanes_mph`, `strep_tb` is built into the `medicaldata` package.

`medicaldata` is the first package this book uses that doesn't come with the tidyverse, so it almost certainly isn't on your machine yet. It has to be installed before it can be loaded. Run `install.packages('medicaldata')` once in the console. Once it's installed, `library(medicaldata)` below works for this session, like every other package we load.

```{r}
#| label: setup
#| message: false

library(readr)
library(dplyr)
library(here)
library(medicaldata)

nhanes <- read_csv(here('data', 'mph_nhanes_20241107.csv'))
dictionary <- read_csv(here('data', 'mph_nhanes_20241107_dictionary.csv'))
```

There's no `library(ggplot2)`, because this chapter draws no plots and there's no reason to load a package we aren't going to use. And `dictionary` is read in even though no code below refers to it, because it's the thing we reach for the moment we're unsure what a column means.

```{r}
#| label: setup-hidden
#| include: false
options(width = 76)
```

```{r}
#| label: recode

nhanes <- nhanes |>
  mutate(
    gender = factor(gender,
      levels = c(1, 2),
      labels = c('Male', 'Female')),
    eth = factor(eth,
      levels = c(1, 2, 3, 4, 5),
      labels = c('Mexican American', 'Other Hispanic', 'White', 'Black', 'Others')),
    edu = factor(edu,
      levels = c(1, 2, 3),
      labels = c('Less than high school', 'High school graduate', 'Above high school')),
    smoking = factor(smoking,
      levels = c('Never', 'Previous', 'Current')),
    phq.c = factor(phq.c,
      levels = c('0 Normal', '1 Mild symptoms', '2 Possible depression'),
      labels = c('Normal', 'Mild symptoms', 'Possible depression'),
      ordered = TRUE),
    blm.cvd = factor(blm.cvd,
      levels = c(0, 1),
      labels = c('No', 'Yes'))
  )
```

This is the same recode we introduced in Chapter 3. See Chapter 3 for the reasoning behind each variable. We repeat it here, unchanged, so this chapter stands on its own too.

## Point estimate and standard error, by hand

A mean handed to a health board with no sense of how precise it is invites a decision the data may not support. A point estimate is our 'best guess' from the data, most commonly the mean, and on its own it says nothing about how precise it is. The standard error (SE) attaches that precision, calculated from two quantities already familiar from Chapter 3: the standard deviation, `sd()`, and the sample size, `n()`. Let's apply this to sedentary time (`sed`):

```{r}
#| label: se-by-hand
sed_stats <- nhanes |>
  filter(!is.na(sed)) |>
  summarise(
    mean_sed = mean(sed),
    sd_sed = sd(sed),
    n_sed = n()
  ) |>
  mutate(se_sed = sd_sed / sqrt(n_sed))

sed_stats
```

`sqrt()` is base R's square root function. The standard error of a mean is the standard deviation divided by the square root of the sample size, the algebra behind the lecture's precision argument: as `n` grows, the denominator grows and the standard error shrinks.

`sed` was shown in Chapter 3 to be right-skewed, with a long tail of participants reporting many sedentary hours. That skew doesn't disqualify it from a CI on its *mean*, because of the central limit theorem. Picture drawing sample after sample from the same population and working out the mean each time: those means vary, and the pattern they make is the *sampling distribution* of the mean. The standard error is its spread. The theorem says that as the sample size grows, the sampling distribution gets closer to a bell-shaped normal distribution, whatever shape the data themselves have.

## From standard error to a CI

A 95% CI for a mean is the point estimate plus or minus 1.96 standard errors. 1.96 is the value that cuts off the middle 95% of a normal distribution:

```{r}
#| label: ci-by-hand
sed_stats <- sed_stats |>
  mutate(
    ci_lower = mean_sed - 1.96 * se_sed,
    ci_upper = mean_sed + 1.96 * se_sed
  )

sed_stats
```

R has a function that does this directly. `t.test()` expects a single numeric vector, not a data frame, so we pull `sed` out of `nhanes` with `$`. `nhanes$sed` reads as 'the `sed` column, inside `nhanes`', and it returns that column on its own, as a plain vector of numbers, not a data frame containing it. That's different from `select(sed)`, which keeps a one-column data frame instead of pulling the column out. `t.test()` then returns the mean and its 95% CI without any of the arithmetic above:

```{r}
#| label: ci-ttest
t.test(nhanes$sed)
```

`t.test()` also reports a test statistic, degrees of freedom, and a p-value. These belong to a hypothesis test, which is Chapter 5's subject. For now, look only at the CI, and check that it closely matches the CI we calculated by hand above.

One thing `t.test()` does without telling us is ignore all missing values. We passed it `nhanes$sed` whole, missing values and all, with no `filter()` and no `na.rm = TRUE`, and it ran anyway. It drops missing values itself, before calculating. The evidence is in the output: the degrees of freedom are `r format(sum(!is.na(nhanes$sed)) - 1, big.mark = ',')`, one less than the `r format(sum(!is.na(nhanes$sed)), big.mark = ',')` participants we filtered down to by hand, which is why the two CIs agree at all. So whenever we run a `t.test()` in R, we should count the missing values ourselves.

The two CIs are close but not identical, because `t.test()` doesn't use a fixed multiplier of 1.96. It uses a value from the t-distribution, which is usually a better match to the actual sampling distribution, since it depends on the sample size through its degrees of freedom (`n - 1`). For a sample size this large, the normal and t-distributions are very similar and shouldn't make a material difference. In practice, we'd use `t.test()` instead of working this out by hand every time, but the formula above is what it calculates.

### Saying what the CI means, and what it doesn't

We now have a CI. Writing it down correctly is a separate skill from calculating it, and it's the one the lecture spent time on, contrasting the right sentence with the wrong one. Here is the right one, for the CI we just built:

> Mean sedentary time was `r round(sed_stats$mean_sed, 2)` hours/day (95%
> CI: `r round(sed_stats$ci_lower, 2)` to `r round(sed_stats$ci_upper, 2)`
> hours/day).

And here is the wrong one, which is the single most common error in this part of the course:

> There is a 95% probability that the true mean sedentary time lies between
> `r round(sed_stats$ci_lower, 2)` and `r round(sed_stats$ci_upper, 2)`
> hours/day.

The true mean sedentary time in the population is a fixed number. We don't know it, but it isn't moving around, and it doesn't have a 95% chance of being anywhere. What moves is the *CI*. Draw a different sample and we get a different mean, a different standard error, and so a different pair of numbers. The 95% belongs to that procedure: of all the CIs we could build this way, 95% of them would contain the true value. This particular CI either contains it or it doesn't, and we have no way of knowing which.

In practice the fix is easy, because the correct sentence is the shorter one. State the estimate, put the CI after it, and stop. The moment we write 'probability', 'chance', or 'there's a 95% likelihood the true value is', we have said something about the parameter that the CI doesn't support. Every CI in the rest of this book gets reported the first way, and so should yours.

## A CI for a proportion

`strep_tb`, from the `medicaldata` package, records the 1948 MRC streptomycin trial: a randomised controlled trial of streptomycin against tuberculosis in 107 patients, generally considered the first modern randomised controlled trial. Each patient was assigned to the `Streptomycin` or `Control` arm and followed up. Patients in the control arm received bed rest alone, not a placebo. `improved` records whether their chest X-ray had improved at six months. As with any new dataset, we look at what came in before writing code against it:

```{r}
#| label: glimpse-strep
glimpse(strep_tb)
```

`arm` and `gender` are already factors. `improved`, the outcome we care about here, is **logical**: `TRUE` or `FALSE`, not a factor. R stores `TRUE` and `FALSE` as 1 and 0, which will matter again shortly.

Below we will look at the proportion of the outcome. `table()`, already met in Chapter 3, counts how many patients fall into each category of `improved`, and `prop.table()` converts those counts to proportions:

```{r}
#| label: prop-table-improved
table(strep_tb$improved)
prop.table(table(strep_tb$improved))
```

### The standard error of a proportion, by hand

A proportion is an estimate like any other, so it has a standard error too. The formula is different from the one for a mean, because a proportion carries its own variability: once we know p, we know how spread out a yes/no variable is, so there's no separate SD to measure. The standard error of a proportion is the square root of p × (1 − p) ÷ n.

`sum()` of a logical vector counts how many values are `TRUE`, since R stores `TRUE` as 1 and `FALSE` as 0. So `sum(strep_tb$improved)` is the number of patients who improved, out of `nrow(strep_tb)` in total.

```{r}
#| label: se-proportion
tb_stats <- strep_tb |>
  summarise(
    n_improved = sum(improved),
    n_total = n(),
    p_improved = n_improved / n_total,
    se_improved = sqrt(p_improved * (1 - p_improved) / n_total)
  )

tb_stats
```

The CI is then built the same way it was for a mean: the estimate, plus or minus 1.96 standard errors.

```{r}
#| label: ci-proportion-hand
tb_stats$p_improved - 1.96 * tb_stats$se_improved
tb_stats$p_improved + 1.96 * tb_stats$se_improved
```

### The same CI from `prop.test()`

`prop.test()` returns a CI for a proportion directly, given a count of successes (`x`) and a total (`n`), with none of the arithmetic above:

```{r}
#| label: prop-test-single
prop.test(x = tb_stats$n_improved, n = tb_stats$n_total)
```

The two CIs are close but not identical, and for the same kind of reason the hand-built and `t.test()` CIs differed earlier. The ± 1.96 version above is the textbook approximation, and it works well when n is large and p isn't near 0 or 1. `prop.test()` uses a better method that behaves more sensibly in small samples and never runs past 0 or 1, which the approximation happily will if pushed too far. With 107 patients and a proportion near a half, both are perfectly reasonable and they agree to within a small margin of error.

Giving `prop.test()` the count and total directly, instead of a cross-tabulated table, avoids any doubt about which of the two categories we're counting. A table of `improved` alone would still show both proportions, but a table fed straight into `prop.test()` reports the proportion for only whichever category comes first, which is not necessarily the one we want.

By default, `prop.test()` also tests whether this proportion differs from 50%. That default comparison isn't of interest here, and it isn't the subject of this chapter. The 95% CI it reports alongside is what we want.

## A CI for a difference in means

`t.test()` also compares two groups directly, using **formula syntax**. `chol ~ gender` reads as '`chol`, modelled by `gender`'. R splits `chol` into the groups defined by `gender` and compares them:

```{r}
#| label: ci-diff-means
t.test(chol ~ gender, data = nhanes |> filter(!is.na(gender)))
```

Notice there's no `$` anywhere in that call, even though `chol` and `gender` both are only found inside `nhanes`. The `data = ` argument tells `t.test()` which data frame to search for names like `chol` and `gender`, so it looks them up there itself instead of us pulling each one out by hand first with `$`. `$` is still the right tool for getting a single column on its own, the way `nhanes$sed` was above; a formula's `data = ` argument does that lookup for every variable named in the formula at once.

This is the same pairing we showed descriptively in Chapter 3's boxplot and Table 1: cholesterol by gender. Now it's attached to a CI for the *difference* in means between the two groups. By default, `t.test()` doesn't assume the two groups have equal variance, that is, the same spread (the variance is the SD squared). This version is called *Welch's t-test*. That's a sensible default and not something we need to adjust here.

### Which way round is the difference?

Both ends of that CI are negative, and a negative difference in cholesterol isn't obviously meaningful until we know what has been subtracted from what. R tells us, on the line beginning 'alternative hypothesis': *true difference in means between group Male and group Female*. So it has taken the men's mean and subtracted the women's.

That order isn't R's choice, and it isn't alphabetical. It's the order the factor's levels are in, which we set ourselves in the setup chunk:

```{r}
#| label: levels-gender-direction
levels(nhanes$gender)
```

`'Male'` is first, so men are group 1 and women are group 2, and the difference runs first minus second. A negative CI therefore says the first group is *lower*: mean cholesterol is lower in men than in women, by somewhere between 0.15 and 0.19 mmol/L.

It is useful to check the alternative-hypothesis line before reading any two-group CI. If we'd rather write the comparison the other way round, which is often more natural to read, we flip the whole thing together. 'Cholesterol was 0.17 mmol/L higher in women than in men (95% CI: 0.15 to 0.19 mmol/L)' says the same as R's output, with every sign reversed. What we must never do is flip the words and leave the numbers, or flip one end of the CI and not the other. The direction of the estimate and the direction of the CI have to agree.

## A CI for a difference in proportions

The same idea extends to two proportions: how many improved in each trial arm. `group_by()` and `summarise()`, already familiar from Chapter 3, give us the count and the number improved in each arm:

```{r}
#| label: arm-summary
strep_arm_summary <- strep_tb |>
  group_by(arm) |>
  summarise(
    n_arm = n(),
    n_improved = sum(improved),
    prop_improved = n_improved / n_arm
  )

strep_arm_summary
```

`prop_improved` here is what `mean(improved)` would give directly. `mean()` of a logical vector is a shortcut worth knowing: R treats `TRUE` as 1 and `FALSE` as 0, so the mean of a column of `TRUE`/`FALSE` values is the proportion that are `TRUE`.

If we pass the two counts and the two totals to `prop.test()`, instead of a cross-tabulated table, it compares the two proportions unambiguously. `prop 1` and `prop 2` below correspond, in order, to the two rows of `strep_arm_summary` above:

```{r}
#| label: ci-diff-props
prop.test(x = strep_arm_summary$n_improved, n = strep_arm_summary$n_arm)
```

```{r}
#| label: ci-diff-props-numbers
#| include: false
diff_props_test <- prop.test(x = strep_arm_summary$n_improved, n = strep_arm_summary$n_arm)
```

The CI is for the difference in the proportion improved between the two arms. Before reading it, check the direction, as we did for the difference in means above. The same rule applies: R takes the first group and subtracts the second, and the order comes from the factor's levels.

```{r}
#| label: levels-arm-direction
levels(strep_tb$arm)
```

`'Streptomycin'` is first, so `prop 1` is the streptomycin arm and `prop 2` is the control arm, and the difference runs streptomycin minus control. Both ends of the CI are positive, which therefore says the streptomycin arm improved *more*. Had the two levels been the other way round, every number in that CI would have carried a minus sign and meant the same thing. Chapter 5 puts this same factor the other way round on purpose, which is why checking the levels here first matters.

The heading says 'with continuity correction': `prop.test()` applies a small adjustment by default that makes the CI slightly wider. Chapter 5 explains what it does, so leave it for now. The p-value printed alongside belongs to a hypothesis test, which is still Chapter 5's subject.

Because `strep_tb` is a randomised trial, not an observational study, a difference this large and this precisely estimated supports a causal reading in a way that an association in `nhanes_mph` would not. We can quite safely conclude, from the findings, that streptomycin improved the proportion of patients with X-ray improvement by about `r round((strep_arm_summary$prop_improved[1] - strep_arm_summary$prop_improved[2]) * 100)` percentage points (95% CI: `r round(diff_props_test$conf.int[1] * 100)` to `r round(diff_props_test$conf.int[2] * 100)` percentage points), compared with bed rest alone. We say percentage points rather than percent here: percent would describe how much bigger one proportion is relative to the other, not the size of the gap between them, which is what this CI estimates.

## Width and sample size, without programming

The lecture's rule was that a CI narrows as sample size grows. We can show this directly, by computing the same CI on the full sample and on two smaller subsets of it. `slice_sample()`, from `dplyr`, keeps a random fraction of the rows. `set.seed()`, from base R, fixes the sequence of 'random' numbers R uses, so the same subset gets drawn every time this code is run, instead of a different one each time.

The full sample. This time we save the result into an object with `<-` as well as printing it, because we want to reuse its numbers in a moment:

```{r}
#| label: width-full
tt_full <- t.test(nhanes$chol)
tt_full
```

A 50% subset:

```{r}
#| label: width-50
set.seed(1)
chol_50 <- nhanes |>
  filter(!is.na(chol)) |>
  slice_sample(prop = 0.5)

tt_50 <- t.test(chol_50$chol)
tt_50
```

A 10% subset:

```{r}
#| label: width-10
set.seed(1)
chol_10 <- nhanes |>
  filter(!is.na(chol)) |>
  slice_sample(prop = 0.1)

tt_10 <- t.test(chol_10$chol)
tt_10
```

The point estimates are similar across all three, because all three are samples from the same population. But the CI widens as the sample size drops.

That's hard to see across three blocks of output, so let's put the three CIs in one small table. A `t.test()` result isn't a data frame, but `$` still works on it, pulling out one named piece at a time, as it pulls one column out of a data frame. `tt_full$conf.int` is the pair of numbers at the ends of the CI, and `[1]` and `[2]` take the lower and the upper one, the same square-bracket indexing we used in Chapter 2. `data.frame()`, from Chapter 1, builds the table out of them, and `mutate()` adds the column we care about:

```{r}
#| label: width-table
widths <- data.frame(
  sample = c('Full', '50% subset', '10% subset'),
  n = c(sum(!is.na(nhanes$chol)), nrow(chol_50), nrow(chol_10)),
  lower = c(tt_full$conf.int[1], tt_50$conf.int[1], tt_10$conf.int[1]),
  upper = c(tt_full$conf.int[2], tt_50$conf.int[2], tt_10$conf.int[2])
) |>
  mutate(width = upper - lower)

widths
```

Cut the sample in half and the CI gets about 40% wider; cut it to a tenth and it's roughly three times wider than we started. Ten times less data did not make the CI ten times wider, it made it about three times wider, and three is roughly the square root of ten. That's the SE = SD/√n formula showing up in an actual result. Precision improves with the square root of the sample size, not with the sample size itself. That's why doubling a study's size buys much less than it feels like it should.

## Sample size and the margin of error

The same logic runs in reverse. Instead of asking how precise an estimate from a given sample size will be, we can choose the precision we want first and work out the sample size that delivers it. The lecture called that precision the **margin of error**: the ± part of estimate ± 1.96 × SE, which is half the CI's width.

For a proportion, the SE is the square root of p × (1 − p) ÷ n, so the margin of error m is 1.96 times that. Rearranged to give n instead:

n = 1.96² × p × (1 − p) ÷ m²

Here p is the proportion we expect to find and m is the margin we want. A survey expecting about half of people to answer yes, and wanting a margin of 3 percentage points either side:

```{r}
#| label: n-moe-prop
1.96^2 * 0.5 * (1 - 0.5) / 0.03^2
```

`^` raises a number to a power, as in Chapter 1's BMI exercise, so `0.03^2` is 0.03 squared. We can't recruit part of a person, so a sample size always rounds up: 1,068 people. Now halve the margin, to 1.5 percentage points:

```{r}
#| label: n-moe-prop-half
1.96^2 * 0.5 * (1 - 0.5) / 0.015^2
```

About four times as many people. m is squared in the formula, so halving it quadruples n. That's the square-root relationship from the widths table above, read the other way round.

A mean works the same way, with the SD taking the place of p × (1 − p): the margin of error is 1.96 × SD ÷ √n, so n = (1.96 × SD ÷ m)². We need an SD to plug in, and a planned study would take it from earlier data. Here we take it from `chol`:

```{r}
#| label: n-moe-mean
chol_sd <- nhanes |>
  filter(!is.na(chol)) |>
  summarise(sd_chol = sd(chol)) |>
  pull(sd_chol)

chol_sd
(1.96 * chol_sd / 0.05)^2
```

`pull()`, from `dplyr`, extracts a single column as a plain vector instead of a one-column tibble. We need that because the formula wants a single number for the SD, not a table containing one. Estimating mean cholesterol to within 0.05 mmol/L either side takes about `r format(ceiling((1.96 * chol_sd / 0.05)^2), big.mark = ',')` people. A noisier measurement, with a larger SD, would need more.

This sizes a study to estimate one quantity to a chosen precision, which is what a descriptive study or a survey needs. Sizing a trial to *detect a difference* between two groups rests on a different idea, statistical power, which Chapter 5 introduces alongside the hypothesis tests it depends on.

::: {.callout-note}
## Try it yourself: the margin you ask for

Re-run the cholesterol calculation with a margin of 0.1 mmol/L, then 0.025 mmol/L. Watch what happens to n each time, and say in one sentence what it means for a researcher who wants a very precise estimate.
:::

## Worked example: a CI for age

A mean on its own says nothing about how much confidence we should place in it. So how precisely can we pin down the average age of participants in this sample?

```{r}
#| label: worked-age-hand
age_stats <- nhanes |>
  filter(!is.na(age)) |>
  summarise(
    mean_age = mean(age),
    sd_age = sd(age),
    n_age = n()
  ) |>
  mutate(
    se_age = sd_age / sqrt(n_age),
    ci_lower = mean_age - 1.96 * se_age,
    ci_upper = mean_age + 1.96 * se_age
  )

age_stats
```

```{r}
#| label: worked-age-ttest
t.test(nhanes$age)
```

The hand-calculated CI and the one from `t.test()` agree closely. Written as it would appear in a results section: the mean age of participants was `r round(age_stats$mean_age, 1)` years (95% CI: `r round(age_stats$ci_lower, 1)` to `r round(age_stats$ci_upper, 1)` years). This is a statement about the precision of the *estimate* of the mean, not a description of how spread out individual ages are. That's what the SD is for.

## Exercises

**Exercise 1** (ILOs 3, 4, 5). Calculate the mean and standard error of total cholesterol (`chol`), construct a 95% CI by hand, and confirm it with `t.test()`. Report the result the way a results section would state it, in one sentence.

Then use `t.test()` with formula syntax to obtain a 95% CI for the *difference* in mean cholesterol between smoking groups, comparing current smokers with those who have never smoked. You'll need `filter()` to restrict `smoking` to those two categories first, since `t.test()` compares exactly two groups. Write one sentence interpreting that CI, and one sentence saying how it differs in meaning from the CI you built in the first half of this exercise.

**Exercise 2** (ILOs 3, 4, 5). In `strep_tb`, find the proportion of patients with baseline cavitation present (`baseline_cavitation`) and calculate its standard error by hand, then build a 95% CI from it and check the result against `prop.test()`. Then construct a 95% CI for the difference in the proportion improved between men and women (`gender`). Write one interpretive sentence for each CI.

**Exercise 3** (ILO 6). Repeat the width exploration above (full sample, 50% subset, 10% subset) using `sed` instead of `chol`, and collect the three CIs into one table the way the chapter did, so their widths sit side by side. Describe what changes and what stays about the same as the sample size drops.

**Exercise 4** (ILO 6). A health board wants to estimate the proportion of adults who currently smoke, and expects it to be about 20%. Find the sample size needed for a 95% CI with a margin of error of 2 percentage points, then 1 percentage point. Then find the sample size needed to estimate mean sedentary time (`sed`) to within 0.5 hours/day, using the SD of `sed` in `nhanes`. Write one sentence explaining why the two smoking sample sizes differ so much.

## Writing exercise

Using the cholesterol CI from Exercise 1 and the `strep_tb` arm-improvement CI from earlier in this chapter, write one sentence for each, in the same reporting style. Use the correct units, and don't describe the CI as a probability statement about where the true value falls.

::: {.callout-note collapse='true'}
## Solutions

**Exercise 1**

```{r}
#| label: sol-ex1
chol_stats <- nhanes |>
  filter(!is.na(chol)) |>
  summarise(
    mean_chol = mean(chol),
    sd_chol = sd(chol),
    n_chol = n()
  ) |>
  mutate(
    se_chol = sd_chol / sqrt(n_chol),
    ci_lower = mean_chol - 1.96 * se_chol,
    ci_upper = mean_chol + 1.96 * se_chol
  )

chol_stats

t.test(nhanes$chol)
```

Mean total cholesterol in this sample was `r round(chol_stats$mean_chol, 2)` mmol/L (95% CI: `r round(chol_stats$ci_lower, 2)` to `r round(chol_stats$ci_upper, 2)` mmol/L).

For the difference between the two smoking groups:

```{r}
#| label: sol-ex1-diff
chol_smoking <- nhanes |>
  filter(smoking %in% c('Never', 'Current'))

t.test(chol ~ smoking, data = chol_smoking)
```

```{r}
#| label: sol-ex1-diff-numbers
#| include: false
tt_smoking <- t.test(chol ~ smoking, data = chol_smoking)
```

`smoking` still has three levels after the `filter()`, but only two of them have any rows left, and `t.test()` rebuilds the factor from the data it's given, so the empty third group drops out on its own.

Check the direction before reading the CI, as above. `'Never'` comes first in `smoking`'s levels, so R has subtracted current smokers from never smokers, and the difference is never minus current.

Mean total cholesterol was `r round(tt_smoking$estimate[1] - tt_smoking$estimate[2], 3)` mmol/L in never smokers compared with current smokers (95% CI: `r round(tt_smoking$conf.int[1], 3)` to `r round(tt_smoking$conf.int[2], 3)` mmol/L), so a little lower on average in those who had never smoked. Notice that the estimate sits inside its own CI. Always check that: if it doesn't, something is wrong, likely a sign has been flipped somewhere. The CI includes zero, so these data are consistent with no difference between the two groups, though they don't rule out a small one either.

The two CIs mean different things. The first is about a single quantity, mean cholesterol, and says how precisely this sample pins it down. The second is about a *comparison* between two groups, and the value that matters in it is zero: a CI containing zero is one where the data can't establish which group is higher. From Chapter 5 onward, almost every CI in this course is of the second kind.

**Exercise 2**

```{r}
#| label: sol-ex2
cav_stats <- strep_tb |>
  summarise(
    n_cav = sum(baseline_cavitation == 'yes'),
    n_total = n(),
    p_cav = n_cav / n_total,
    se_cav = sqrt(p_cav * (1 - p_cav) / n_total)
  )

cav_stats

cav_stats$p_cav - 1.96 * cav_stats$se_cav
cav_stats$p_cav + 1.96 * cav_stats$se_cav

cav_test <- prop.test(x = cav_stats$n_cav, n = cav_stats$n_total)
cav_test

gender_summary <- strep_tb |>
  group_by(gender) |>
  summarise(n_gender = n(), n_improved = sum(improved))
gender_summary

gender_test <- prop.test(x = gender_summary$n_improved, n = gender_summary$n_gender)
gender_test
```

The hand-built CI and the one from `prop.test()` agree closely, as they did for the proportion improved earlier in the chapter.

Just over half of patients (`r round(cav_stats$p_cav * 100)`%) had cavitation present on their baseline chest X-ray (95% CI: `r round(cav_test$conf.int[1] * 100)`% to `r round(cav_test$conf.int[2] * 100)`%). The CI includes 50%, so these data don't rule out the true proportion being exactly one half. The CI for the difference in improvement between women (`prop 1`) and men (`prop 2`) runs from `r round(gender_test$conf.int[1], 2)` to `r round(gender_test$conf.int[2], 2)`: it includes zero, consistent with little or no difference in improvement by gender in this trial.

**Exercise 3**

```{r}
#| label: sol-ex3
tt_sed_full <- t.test(nhanes$sed)

set.seed(1)
sed_50 <- nhanes |>
  filter(!is.na(sed)) |>
  slice_sample(prop = 0.5)
tt_sed_50 <- t.test(sed_50$sed)

set.seed(1)
sed_10 <- nhanes |>
  filter(!is.na(sed)) |>
  slice_sample(prop = 0.1)
tt_sed_10 <- t.test(sed_10$sed)

data.frame(
  sample = c('Full', '50% subset', '10% subset'),
  n = c(sum(!is.na(nhanes$sed)), nrow(sed_50), nrow(sed_10)),
  lower = c(tt_sed_full$conf.int[1], tt_sed_50$conf.int[1],
            tt_sed_10$conf.int[1]),
  upper = c(tt_sed_full$conf.int[2], tt_sed_50$conf.int[2],
            tt_sed_10$conf.int[2])
) |>
  mutate(width = upper - lower)
```

As with `chol`, the point estimate is broadly similar across all three samples, but the CI widens substantially in the 10% subset, and by about the same factor of three. The same pattern holds for a skewed variable as for a symmetric one, because it's the *sample size*, not the shape of the raw data, that governs the width of the CI on the mean.

**Exercise 4**

```{r}
#| label: sol-ex4
1.96^2 * 0.2 * (1 - 0.2) / 0.02^2
1.96^2 * 0.2 * (1 - 0.2) / 0.01^2

sed_sd <- nhanes |>
  filter(!is.na(sed)) |>
  summarise(sd_sed = sd(sed)) |>
  pull(sd_sed)

(1.96 * sed_sd / 0.5)^2
```

Rounded up, a margin of 2 percentage points needs 1,537 adults, and a margin of 1 percentage point needs 6,147, four times as many. The margin is squared in the formula, so halving it quadruples the sample size: precision is bought with the square root of n. Estimating mean sedentary time to within 0.5 hours/day needs about `r format(ceiling((1.96 * sd(nhanes$sed, na.rm = TRUE) / 0.5)^2), big.mark = ',')` people. `sed`'s SD, about `r round(sd(nhanes$sed, na.rm = TRUE), 1)` hours/day, is large next to a margin of half an hour, and the SD enters the formula squared.

**Writing exercise (sample answer)**

Mean total cholesterol was `r round(chol_stats$mean_chol, 2)` mmol/L (95% CI: `r round(chol_stats$ci_lower, 2)` to `r round(chol_stats$ci_upper, 2)` mmol/L). In the streptomycin trial, the proportion of patients improved was about `r round((strep_arm_summary$prop_improved[1] - strep_arm_summary$prop_improved[2]) * 100)` percentage points higher in the streptomycin arm than in the control arm (95% CI: `r round(diff_props_test$conf.int[1] * 100)` to `r round(diff_props_test$conf.int[2] * 100)` percentage points). The CI sits entirely above zero, so the data are compatible only with more improvement in the streptomycin arm than in the control arm.
:::
