# Hypothesis tests and measures of association (Week 9) {#sec-09-comparison}

A confidence interval, or CI, says how precise an estimate is. It doesn't say whether the effect behind it is distinguishable from none at all: two groups' cholesterol means could sit half a millimole apart with CIs that overlap, and whether that gap is a real difference or ordinary sampling noise is the question Chapter 4 leaves open. A hypothesis test closes it, and gets reported alongside the CI, not in place of it. Binary outcomes raise a version of the same question the descriptive statistics from Chapter 3 can't answer on their own: is a health outcome more common in one group than another, and by how much? The risk ratio and the odds ratio from epidemiology measure that, and here we attach a CI and a p-value to each.

Every test in this chapter runs on the same logic underneath: a null hypothesis, an alternative, a test statistic, and a p-value stating how surprising the observed data would be if the null were true. Which test applies depends on the shape of the data in front of us, one measurement per person or two, one group or several, continuous or count, and telling those apart correctly matters as much as running the test itself. A test's decision can also be wrong in either direction, and the rate at which it misses a real effect is what sizes a trial before anyone is recruited. Reported together, an effect size, its 95% CI, and a p-value make up the standard this book writes results in from here on.

## Chapter intended learning outcomes

By the end of this week you will be able to:

1. Formulate null and alternative hypotheses for a research question
2. Explain what a p-value is and is not, and define Type I and Type II error and the problem multiple testing creates, and calculate a sample size for a given power with `power.t.test()` and `power.prop.test()`
3. Describe the relationship between a confidence interval and a hypothesis test
4. Select, interpret, and report an appropriate test given the outcome type and the study design (one-sample, paired, or independent), using `t.test()`, `wilcox.test()`, `chisq.test()`, `fisher.test()`, and `epitools::riskratio()` and `epitools::oddsratio()`

## Setup

As in Chapters 3 and 4, we read the data with `read_csv()` and turn the six coded variables into factors against the dictionary. We also bring back `strep_tb`, from the `medicaldata` package, for the categorical half of this chapter.

This chapter adds one new package, `epitools`. Install it with `install.packages('epitools')` in the console, the same way as any other package, if it isn't on your machine already. Once it's installed, `library(epitools)` below makes its functions available for this session.

```{r}
#| label: setup
#| message: false

library(readr)
library(dplyr)
library(ggplot2)
library(gtsummary)
library(here)
library(medicaldata)
library(epitools)

nhanes <- read_csv(here('data', 'mph_nhanes_20241107.csv'))
dictionary <- read_csv(here('data', 'mph_nhanes_20241107_dictionary.csv'))
```

```{r}
#| label: setup-hidden
#| include: false
options(width = 76)
source(here('R', 'gtsummary-knitprint.R'))
```

```{r}
#| label: recode

nhanes <- nhanes |>
  mutate(
    gender = factor(gender,
      levels = c(1, 2),
      labels = c('Male', 'Female')),
    eth = factor(eth,
      levels = c(1, 2, 3, 4, 5),
      labels = c('Mexican American', 'Other Hispanic', 'White', 'Black', 'Others')),
    edu = factor(edu,
      levels = c(1, 2, 3),
      labels = c('Less than high school', 'High school graduate', 'Above high school')),
    smoking = factor(smoking,
      levels = c('Never', 'Previous', 'Current')),
    phq.c = factor(phq.c,
      levels = c('0 Normal', '1 Mild symptoms', '2 Possible depression'),
      labels = c('Normal', 'Mild symptoms', 'Possible depression'),
      ordered = TRUE),
    blm.cvd = factor(blm.cvd,
      levels = c(0, 1),
      labels = c('No', 'Yes'))
  )
```

This is the same recode we introduced in Chapters 3 and 4. See Chapter 3 for the reasoning behind each variable.

## The logic of a test

A classical hypothesis test starts from two competing statements about the **population**. The **null hypothesis**, written H0, says there's nothing going on: no difference between groups, or no association between variables. The exact wording depends on what is being tested, but it always describes the population, never the sample we happen to have measured. The **alternative hypothesis**, H1, says there is a difference. A one-sided alternative hypothesis, specifying the direction of the difference in advance, exists too, but is rarely used in practice; every test in this chapter uses the two-sided form. A test never proves the null hypothesis true. It only asks whether the pattern shown from the sample is surprising enough to reject the null hypothesis.

A test turns the sample data into a single number, a **test statistic**, that measures how far the data sit from what the null hypothesis predicts under the sampling distribution. R then converts that statistic into a **p-value**: the probability of seeing that test statistic at least this extreme, if the null hypothesis were true. A p-value is not the probability that the null hypothesis is true, and it's not the probability the result happened by chance. A small p-value, conventionally below 0.05, means the data would be unusual under the null hypothesis, so we reject it in favour of the alternative. That 0.05 cut-off is the **significance level**, and a result with a p-value below it is called **statistically significant**. A p-value alone says nothing about how large an effect is, only about how surprising it is.

A CI and a hypothesis test carry the same information, read two different ways. If a 95% CI for a difference excludes the null value, 0 for a difference, 1 for a risk ratio or odds ratio, the matching test rejects the null hypothesis at the 5% level, and vice versa. From here on, every test in this book states its null and alternative hypothesis in words before running the test, always as a statement about the population, never about the sample or the particular study, and every result is reported the way the interactive lecture set out: an effect size, a CI, and a p-value, together.

## One-sample t-test: total cholesterol against a clinical benchmark

Guideline bodies often set a single reference value for a measurement: a 'desirable' total cholesterol of under 5.2 mmol/L is a standard clinical benchmark. That specific figure is the American guideline; UK guidance instead sets the desirable level at 5.0 mmol/L. Either is a reasonable fixed value to test a sample's mean against, and this chapter works with 5.2 mmol/L throughout. A one-sample t-test asks whether the population's mean differs from a single fixed value like that one, instead of from another group's mean.

**Null hypothesis.** Mean total cholesterol in the population is 5.2 mmol/L.

**Alternative hypothesis.** Mean total cholesterol in the population is not 5.2 mmol/L.

`t.test()` runs a one-sample test when we give it one variable and a value to test it against, `mu`:

```{r}
#| label: ttest-onesample-chol
t.test(nhanes$chol, mu = 5.2)
```

R prints a 95% CI for the mean itself, 4.96 to 4.98 mmol/L, not for the quantity the test is about, which is how far that mean sits from the 5.2 mmol/L benchmark. Subtracting 5.2 from both ends turns it into the effect size we want: mean total cholesterol in this sample is about 4.97 mmol/L, 0.23 mmol/L below the 5.2 mmol/L benchmark (95% CI: 0.22 to 0.24 mmol/L below, p < 0.001).

A one-sample t-test assumes the **sampling distribution** of the mean is roughly normally distributed, not that `chol` itself is. We never see a sampling distribution directly, since we only have the one sample, so the shape of the data is the evidence we have about it. The central limit theorem says the sampling distribution of the mean gets closer to normal as the sample grows, whatever shape the data themselves have, and the more skewed the data, the larger a sample that takes. Chapter 3's histogram of `chol` showed a distribution with a mild right skew, not a perfectly symmetric bell shape, but with a sample this large, tens of thousands of participants, the sample is easily large enough for the theorem to apply. So the question a histogram answers here is not 'is `chol` normal?' but 'is this sample large enough, given how skewed `chol` is?', and for `chol` the answer is a comfortable yes. `wilcox.test()` is the **non-parametric** alternative, meaning it doesn't rely on the sampling distribution being normal at all. It's the one to reach for with a smaller sample (typically <30) or a more pronounced skew where that reassurance is less safe.

**Null hypothesis (`wilcox.test()`).** The population median of `chol` is 5.2 mmol/L.

**Alternative hypothesis.** The population median is not 5.2 mmol/L.

`wilcox.test()`, from base R, takes the same `mu` argument for a one-sample test. Instead of a mean, it tests a **median**, using the ranks of how far each value sits from `mu`, so one extreme value can't dominate the result the way it can dominate a mean:

```{r}
#| label: wilcox-onesample-chol
wilcox.test(nhanes$chol, mu = 5.2)
```

`wilcox.test()` returns a p-value and nothing else, no effect size, so the size of the effect has to come from a descriptive statistic that matches what the test is about: a median, not a mean. The sample median of `chol` is 4.89 mmol/L (IQR: 4.22 to 5.61 mmol/L), below the 5.2 mmol/L benchmark, consistent with the p-value far below 0.001. Typical cholesterol in this sample sits below the clinical benchmark whichever way we measure the middle of the distribution.

## Paired t-test: two domains of activity in one person

`nhanes_mph` records physical activity as two separate totals for each participant, both in MET-hours/day: `met_w`, activity at work, and `met_r`, activity outside work, as the dictionary describes it. These are two different domains of activity, measured at the same time, in the same person.

That distinction decides which test applies. The one-sample test above compared one variable against a fixed value. Here there's no fixed value: two measurements to compare, both taken on the same people. That's not the same as comparing two separate groups, an independent-samples test, introduced later in this chapter. Every `met_w` value here has exactly one matching `met_r` value, from the same participant. The right tool is a paired t-test. Instead of comparing the two columns directly, it collapses them into a single column of within-person differences (`met_w` minus `met_r`) and tests whether that difference is, on average, zero.

**Null hypothesis.** In the population, mean work activity and mean outside-work activity are the same: the mean within-person difference is zero.

**Alternative hypothesis.** The mean within-person difference is not zero.

`t.test()` runs a paired test when we set `paired = TRUE`:

```{r}
#| label: paired-ttest
t.test(nhanes$met_w, nhanes$met_r, paired = TRUE)
```

On average, work activity ran about 3.07 MET-hours/day higher than activity outside work (95% CI: 2.93 to 3.21 MET-hours/day, p < 0.001).

A paired t-test also arises in a before-and-after design, where the same measurement is taken at two time points on the same person. That is a different study design from the one here, even though the R code looks identical. There is no repeated measurement over time in this data, and describing the result as though there were would plant a habit that causes real trouble in a longitudinal design later on.

We can also get the paired t-test result by running an ordinary one-sample t-test on the differences directly, testing whether `met_w - met_r` has a mean of zero, the calculation the one-sample section above introduced:

```{r}
#| label: paired-ttest-manual
t.test(nhanes$met_w - nhanes$met_r)
```

The t-statistic, degrees of freedom, and p-value are identical to the paired test above. That's not a coincidence: `paired = TRUE` does this calculation internally, collapsing the two columns into one column of differences first.

## A non-parametric alternative: the Wilcoxon signed-rank test

A paired t-test assumes the sampling distribution of the within-person difference is at least roughly normally distributed, the same kind of claim the one-sample section above explained. Let's look at the data's distribution, this time with a histogram, since the within-person difference is a new variable we haven't looked at yet:

```{r}
#| label: hist-met-diff
ggplot(nhanes |> mutate(met_diff = met_w - met_r), aes(x = met_diff)) +
  geom_histogram(binwidth = 5) +
  labs(x = 'Work minus outside-work activity (MET-hours/day)', y = 'Count')
```

The distribution of differences is heavily skewed, and a large share of participants sit exactly at zero. Many people report no occupational activity at all, so `met_w` and `met_r` are equal for them. The paired t-test may still be fine, because the sample size is large enough for the central limit theorem to apply, but a skew this pronounced is a good reason to check the conclusion against a test that doesn't lean on that theorem at all.

`wilcox.test()` also tests the same question as the paired t-test here, without assuming a normal distribution. It works with the ranks of the differences instead of their average, so one extreme value or an oddly-shaped distribution can't dominate the result the way it can dominate a mean.

**Null hypothesis.** In the population, work activity and outside-work activity are equally likely to be the larger of the pair: there's no systematic difference within the pair.

**Alternative hypothesis.** One is systematically larger than the other.

With `paired = TRUE`, it runs the paired version, known as the Wilcoxon signed-rank test:

```{r}
#| label: wilcox-met
wilcox.test(nhanes$met_w, nhanes$met_r, paired = TRUE)
```

The conclusion doesn't change here: both tests reject the null hypothesis convincingly, each with a p-value far below 0.001, and the paired t-test's own effect size above, work activity running about 3.07 MET-hours/day higher, is still the effect-size figure to report alongside either p-value. With a sample this large, the t-test and the Wilcoxon test agree despite the skew. That won't always be true. With a smaller sample, or a more extreme distribution, the two tests can disagree, and the non-parametric result is usually the one to trust when the normality assumption is in real doubt.

## Independent-samples t-test: cholesterol by gender

Unlike the paired comparison just above, an independent-samples t-test compares two entirely separate groups of people, with no link between a value in one group and a value in the other. `t.test()` compares two independent groups directly. We already used it in Chapter 4 to build a CI for the difference in total cholesterol (`chol`, in mmol/L) between men and women.

**Null hypothesis.** Mean total cholesterol is the same in men and women in the population.

**Alternative hypothesis.** Mean total cholesterol differs between men and women in the population.

```{r}
#| label: ttest-chol-gender
t.test(chol ~ gender, data = nhanes |> filter(!is.na(gender)))
```

This is the same call as Chapter 4's, where we only read the CI. Now we read the rest: a t-statistic, degrees of freedom, and a p-value.

Before trusting a t-test, we check that the two groups aren't wildly different in spread or shape, the way Chapter 3's boxplot of `chol` by `gender` already let us do. This is a different check from the sampling-distribution question the one-sample and paired sections asked: there, the concern was whether one sampling distribution was close to normal; here, with two separate groups, the concern is whether the two groups look comparable enough in spread and shape for a difference in their means to be meaningful.

```{r}
#| label: boxplot-recheck
ggplot(nhanes |> filter(!is.na(gender)), aes(x = gender, y = chol)) +
  geom_boxplot() +
  labs(x = 'Gender', y = 'Total cholesterol (mmol/L)')
```

The two boxplots are similar in spread and shape, just shifted slightly. Nothing here rules out a t-test. If, instead, the shape of the boxplot looks very different between the two groups, a t-test might not be appropriate, since it only compares the sample means. `t.test()` also doesn't assume equal variance between the two groups by default, which protects us even where the spreads differ more than they do here.

Before reading the CI, check the direction, as Chapter 4 set out. The alternative-hypothesis line says *group Male and group Female*, because `'Male'` is the first level of `gender`, so R has printed men minus women and both ends come out negative. Reversing every sign together turns it into the more natural reading: mean total cholesterol was about 0.17 mmol/L higher in women than in men (95% CI: 0.15 to 0.19 mmol/L, p < 0.001). `p < 0.001` is how we write a p-value that R prints as something like `< 2.2e-16`. Convention in public health (or clinical medicine) reports a very small p-value this way instead of typing out the scientific notation.

## A non-parametric alternative: the Mann-Whitney test

The independent-samples t-test has a non-parametric alternative too, the **Mann-Whitney test**, also called the Wilcoxon rank-sum test. It puts every value from both groups into one ranked list, then asks whether one group's values sit systematically higher up that list than the other's. Like the signed-rank test above, it doesn't rely on a normal sampling distribution, so it's the one to reach for when the groups are small or their distributions strongly skewed.

**Null hypothesis.** In the population, neither men's nor women's cholesterol values tend to be higher than the other group's.

**Alternative hypothesis.** One group's values tend to be higher than the other's.

`wilcox.test()` runs the rank-sum version when we give it a formula with a two-group factor, the same way `t.test()` took one:

```{r}
#| label: wilcox-chol-gender
wilcox.test(chol ~ gender, data = nhanes |> filter(!is.na(gender)))
```

As with the one-sample version, the output is a p-value and no effect size. The effect size comes from the matching descriptive statistics, the median and quartiles in each group, which `group_by()` and `summarise()` give us the same way Chapter 3 built them:

```{r}
#| label: median-chol-gender
chol_gender_medians <- nhanes |>
  filter(!is.na(gender), !is.na(chol)) |>
  group_by(gender) |>
  summarise(
    median_chol = median(chol),
    q1_chol = quantile(chol, probs = 0.25),
    q3_chol = quantile(chol, probs = 0.75)
  )

chol_gender_medians
```

```{r}
#| label: wilcox-chol-gender-numbers
#| include: false
mw_p <- wilcox.test(chol ~ gender, data = nhanes |> filter(!is.na(gender)))$p.value
mw_m <- chol_gender_medians |> filter(gender == 'Male')
mw_f <- chol_gender_medians |> filter(gender == 'Female')
```

Median total cholesterol was `r round(mw_f$median_chol, 2)` mmol/L in women (IQR: `r round(mw_f$q1_chol, 2)` to `r round(mw_f$q3_chol, 2)` mmol/L) and `r round(mw_m$median_chol, 2)` mmol/L in men (IQR: `r round(mw_m$q1_chol, 2)` to `r round(mw_m$q3_chol, 2)` mmol/L), `r if (mw_p < 0.001) 'p < 0.001' else paste('p =', round(mw_p, 3))`. That's the same conclusion as the t-test, which is what we'd expect from two large groups with similar spread and shape.

## Chi-squared test: `strep_tb`'s primary result

`strep_tb`'s primary outcome is binary: did chest X-ray improve at six months (`improved`), and does that depend on trial arm (`arm`)? `table()` builds a two-way count, the same way it did in Chapter 3, and `chisq.test()`, from base R, tests whether the two variables are associated.

**Null hypothesis.** The probability of improvement is the same in the population of patients receiving streptomycin and the population of patients receiving control treatment.

**Alternative hypothesis.** The probability of improvement is not the same in the two populations.

```{r}
#| label: table-arm-improved
tab_arm <- table(strep_tb$arm, strep_tb$improved)
tab_arm
```

```{r}
#| label: chisq-arm-improved
chisq.test(tab_arm, correct = FALSE)
```

The chi-squared statistic is large and the p-value is small: the proportion improved is not the same in the two arms. A p-value alone doesn't say by how much; the risk ratio and odds ratio below put a size on that difference.

A chi-squared test also has an assumption to check. Every **expected** count should be reasonably large, conventionally at least 5. `chisq.test()` calculates these expected counts internally and stores them in `$expected`:

```{r}
#| label: chisq-expected
chisq.test(tab_arm, correct = FALSE)$expected
```

All four expected counts are comfortably above 5, so the test is on solid ground here. Expected counts run low typically in a small or very unevenly split sample. When that happens, `fisher.test()`, also from base R, is the alternative: it tests the same association by exact calculation, not by approximation, and doesn't depend on that condition.

`correct = FALSE` switches off a specific adjustment. Left alone on a 2×2 table, `chisq.test()` applies **Yates' continuity correction** by default: it subtracts 0.5 from the absolute difference between each observed and expected count before squaring, a conservative adjustment that compensates for approximating whole-number counts with a smooth curve. It shrinks the test statistic slightly, so it makes the p-value larger.

We've already met it without being told what it was. Chapter 4's `prop.test()` on these same two arms was headed 'with continuity correction' too, and printed a p-value of 0.0003548. The uncorrected test above gives 0.000166: same data, same comparison, two different numbers, purely from this one default.

The correction earns its place when expected counts run low, typically down around 5 to 10. Ours aren't: the smallest expected count is above 25. So we switch it off here, and it's also what makes this p-value match the one `epitools` reports for the same table in the next section.

::: {.callout-note}
## Try it yourself: two tests, one answer

Chapter 4's `prop.test()` compares two proportions; this section's `chisq.test()` tests association in a two-way table. Run `prop.test(tab_arm, correct = FALSE)` and compare its chi-squared statistic and p-value to `chisq.test(tab_arm, correct = FALSE)` above, digit for digit. Write one sentence on what that match tells us about how the two tests relate, for a 2×2 table like this one.

`prop.test()` also prints something `chisq.test()` doesn't: a CI for the difference between the two proportions. Read the two proportions R prints and check which category of `improved` they refer to, the same way Chapter 4 warned to check before feeding a table straight in. Reverse the signs if needed, the way the independent-samples section above reversed its own CI, and write a one-sentence results statement giving the risk difference between the two arms with its 95% CI. Write a second sentence on why that statement is something `chisq.test()` alone could never have given us.
:::

## Fisher's exact test

This is the test to reach for when a chi-squared test's expected-count assumption fails, which in public health work usually means a small trial or a rare outcome. It tests the same null and alternative hypothesis as the chi-squared test above, independence versus dependence between arm and improvement, by an exact calculation instead of an approximation. Let's try `fisher.test()` on this table too, even though nothing here required it. The call looks just like `chisq.test()`:

```{r}
#| label: fisher
fisher.test(tab_arm)
```

The p-value, 0.0002218, is close to the chi-squared test's 0.000166. Both lead to the same conclusion here, a clear association between arm and improvement. `riskratio()` and `oddsratio()`, met below, report a Fisher's exact p-value of their own for the same table, so this number will reappear.

On a larger table, with many categories instead of the 2×2 shape here, `fisher.test()` can become slow, because it enumerates every possible table with the same row and column totals to work out the exact probability. If it runs too slowly or refuses with an error on a bigger table, adding `simulate.p.value = TRUE` switches it to a Monte Carlo estimate instead of the exact calculation, trading a small amount of precision for speed. That isn't needed here.

## Risk ratios and odds ratios, with CIs

The epidemiology half already taught how to calculate a risk ratio and an odds ratio from a 2×2 table like `tab_arm` above, and what each one means. We don't re-derive those formulas here. What's new is attaching a CI and a p-value to numbers we can already compute by hand, and checking that R's version agrees with ours.

`riskratio()` and `oddsratio()`, both from `epitools`, take a 2×2 table and compare its rows, treating the first row as the reference category everything else is measured against.

**Null hypothesis.** In the population, the risk ratio is 1, and the odds ratio is 1: no association between arm and improvement.

**Alternative hypothesis.** The risk ratio, and the odds ratio, differ from 1.

Note the null value sits at 1 for a risk ratio or an odds ratio, not 0. A ratio of 1 means the two groups' risk, or odds, are identical; a difference in means or proportions has its null at 0 instead, because there the null value means no gap, not no ratio.

They expect the table a particular way round, and getting it wrong produces a perfectly ordinary-looking answer, not an error, so check the layout before building one. The **exposure goes in the rows**, with the reference group first. The **outcome goes in the columns**, and the second column is taken as the event: the thing whose risk we're comparing. Our table has `arm` in the rows and `improved` in the columns. `improved` holds `TRUE`/`FALSE` values, and `table()` orders these `FALSE` then `TRUE`, so improvement lands in the second column as needed. Print the table and look at it before passing it on. If the reference row or the event column is the wrong one, these functions still return a risk ratio, an upside-down one.

`levels()`, from base R, lists a factor's categories in their current order, which is what we need to check before comparing rows:

```{r}
#| label: levels-arm
levels(strep_tb$arm)
```

`Streptomycin` is listed first, which would make it the reference, backwards from what we want here. `relevel()`, met in Chapter 3, moves the level named in `ref` to the front, so it becomes the new reference level. There we were only demonstrating it; here it changes the answer:

```{r}
#| label: relevel-arm
arm_control_ref <- relevel(strep_tb$arm, ref = 'Control')
levels(arm_control_ref)
```

`Control` is now first, so it becomes the reference row: the one everything else is compared against.

We rebuild the table with the releveled factor, and pass it to `riskratio()`:

```{r}
#| label: riskratio-arm
tab_arm_releveled <- table(arm_control_ref, strep_tb$improved)
tab_arm_releveled

riskratio(tab_arm_releveled)
```

`oddsratio()` works the same way. We ask for `method = 'wald'`, which uses the same normal-approximation logic behind the ± 1.96 × SE CIs from Chapter 4. This matches the simple cross-product formula from the hand calculation directly:

```{r}
#| label: oddsratio-arm
oddsratio(tab_arm_releveled, method = 'wald')
```

From the table above: 38 of 55 patients improved in the streptomycin arm (risk = 0.69), and 17 of 52 improved in the control arm (risk = 0.33). By hand, risk ratio = 0.69 / 0.33 = 2.11, and odds ratio = (38 × 35) / (17 × 17) = 4.60. Both match what `riskratio()` and `oddsratio()` just reported, with a CI attached to each: a risk ratio of 2.11 (95% CI: 1.38 to 3.24, p < 0.001) and an odds ratio of 4.60 (95% CI: 2.04 to 10.39, p < 0.001). Neither CI comes close to including 1, the null value for both a risk ratio and an odds ratio, consistent with the chi-squared test above.

`riskratio()` and `oddsratio()` each also report three p-values alongside the estimate: a mid-p exact test, Fisher's exact test, and a chi-squared test. All three agree closely here. The chi-squared one is 0.000166, exactly the p-value `chisq.test(tab_arm, correct = FALSE)` gave us above, to every digit printed. That's only because we switched the continuity correction off. Had we left the correction on, `chisq.test()` would have reported 0.0003548, the number Chapter 4's `prop.test()` printed, while `epitools` reported 0.000166. There'd be no way to tell from the output that the two functions had made different default choices.

## Errors, power, and sample size

Every test in this chapter ends in a decision against the 0.05 significance level, and because data are random, that decision can go wrong in two ways. A **Type I error** is rejecting the null hypothesis when it's true: a false positive. The Type I error rate is the significance level itself, so at 0.05, a test of a true null hypothesis still comes out significant 5% of the time. A **Type II error** is failing to reject the null hypothesis when it's false: a false negative. For a fixed sample size, making one of the two errors rarer makes the other more common.

That 5% applies to each test separately. Run twenty independent tests at 0.05, every one of them on a null hypothesis that's true, and on average one comes out significant by chance alone. This is the problem of **multiple testing**: the more tests a study runs, the more of its significant results we'd expect to be false positives.

**Power** is one minus the Type II error rate: the probability that a study detects an effect of a given size, if that effect exists. It rises with a larger effect, a larger sample, or less variable data, and it falls if we choose a stricter significance level.

Chapter 4 sized a study to estimate one quantity to a chosen margin of error. A trial comparing two groups is usually sized on power instead, to give it a good chance of detecting a difference of a given size. `strep_tb` supplies the natural example, since it's a trial with a binary outcome: from the table above, 69% improved in the streptomycin arm and 33% in the control arm.

`power.prop.test()`, from base R, answers that practical question for two proportions. We give it the two proportions, a significance level, and a target power:

```{r}
#| label: power-prop-original
power.prop.test(p1 = 0.33, p2 = 0.69, sig.level = 0.05, power = 0.8)
```

`sig.level = 0.05` sets the Type I error rate, the conventional choice. `power = 0.8` is the conventional target: an 80% chance of detecting the effect if it exists. `n` is the number **per group** required. The gap observed in the original trial was unusually large. A smaller assumed gap needs a correspondingly larger trial:

```{r}
#| label: power-prop-smaller
power.prop.test(p1 = 0.33, p2 = 0.45, sig.level = 0.05, power = 0.8)
```

`power.t.test()` sizes a study around a difference in means instead, using the variability of `chol`. `pull()`, met in Chapter 4, gives us the SD as a single number:

```{r}
#| label: power-t-chol
chol_sd <- nhanes |>
  filter(!is.na(chol)) |>
  summarise(sd_chol = sd(chol)) |>
  pull(sd_chol)

power.t.test(delta = 0.5, sd = chol_sd, sig.level = 0.05, power = 0.8)
```

`delta` is the assumed difference in means the study is being sized to detect. As with the proportions above, a smaller assumed difference requires a larger sample:

```{r}
#| label: power-t-smaller
power.t.test(delta = 0.2, sd = chol_sd, sig.level = 0.05, power = 0.8)
```

Both functions size a study *before* it is run, from an assumed effect. Power calculated after the fact, from the effect a study actually observed, is circular: once the data are in, the estimate and its CI already say what the study could resolve, and 'post-hoc power' only re-expresses the p-value we already have.

::: {.callout-note}
## Try it yourself: assumed effect size

Re-run `power.prop.test()` with `p2` set closer and closer to `p1` (for example 0.40, then 0.35), keeping `power = 0.8`. Watch what happens to `n`, and say in one sentence what it means for a researcher planning a trial to detect a small effect.
:::

## `gtsummary::tbl_summary()` with a test attached

`tbl_summary()`, already met in Chapter 3, can attach a test to a stratified table with `add_p()`, piped on afterwards:

```{r}
#| label: tbl-summary-addp
strep_tb |>
  select(arm, improved) |>
  tbl_summary(by = arm) |>
  add_p()
```

This single call reproduces the count, the percentage, and the p-value we built by hand above, in the one-line form we'll reach for once the manual steps are familiar. The p-value matches to every digit because `add_p()` happens to default to the *uncorrected* chi-squared test here, the same choice we made by hand with `correct = FALSE`.

::: {.callout-note}
## Try it yourself: sample size and significance

Using `slice_sample()` from Chapter 4, take a 10% random subsample of `nhanes` (remember to `set.seed()` first, so the subsample is reproducible) and re-run the `chol ~ gender` t-test from earlier in this chapter on it. Compare the p-value and the CI to the ones from the full sample. Note what changed in one sentence, then explain in a second what it means for judging whether a small p-value reflects a large effect or simply a large sample.
:::

## Worked example: sedentary time by gender

Let's bring the independent-samples t-test together on one more research question: does sedentary time (`sed`, in hours/day) differ between men and women?

```{r}
#| label: worked-boxplot-sed
ggplot(nhanes |> filter(!is.na(gender)), aes(x = gender, y = sed)) +
  geom_boxplot() +
  labs(x = 'Gender', y = 'Sedentary time (hours/day)')
```

```{r}
#| label: worked-ttest-sed
t.test(sed ~ gender, data = nhanes |> filter(!is.na(gender)))
```

The boxplots overlap substantially, with similar spread in each group, so a t-test is reasonable here too. R prints the difference men minus women once again, so the CI on the page runs from -0.49 to -0.03. Put into a sentence, with the whole thing reversed: women reported slightly more sedentary time than men (mean difference about 0.26 hours/day, 95% CI: 0.03 to 0.49 hours/day, p = 0.026). The CI doesn't include zero, so the difference is statistically significant, but it's a small effect, about 15 minutes a day, on a scale of many hours a day.

## Exercises

**Exercise 1** (ILOs 1, 4). State the null and alternative hypotheses for a one-sample test of sedentary time (`sed`) against a reference value of 6 hours/day. Perform a one-sample t-test with `t.test()`, then repeat it with `wilcox.test()`. Report both results in one sentence each, each with its own effect size, and write a third sentence on why the mean and the median of `sed` might point to different conclusions about a typical participant's sedentary time.

**Exercise 2** (ILOs 1, 4). State the null and alternative hypotheses for a paired comparison of `met_w` and `met_r`. Using `filter()`, restrict `nhanes` to participants whose sedentary time (`sed`) is above the sample median. On this subgroup, run both a paired t-test and a Wilcoxon signed-rank test comparing `met_w` and `met_r`. Write one sentence stating whether your conclusion changes in this more sedentary subgroup, and one sentence explaining why `met_w` and `met_r` are still a paired comparison here, not an independent one, even after filtering.

**Exercise 3** (ILOs 1, 4). State the null and alternative hypotheses, then perform an independent-samples t-test comparing age (`age`) between men and women. Check the assumption graphically first with a boxplot. Report the result in a single sentence, and add one more sentence on whether a statistically significant result here is also a practically important one.

**Exercise 4** (ILOs 1, 4). State the null and alternative hypotheses, then construct a contingency table of baseline cavitation (`baseline_cavitation`) against improvement (`improved`) in `strep_tb`. Run a chi-squared test, check the expected counts, and write one sentence interpreting the result.

**Exercise 5** (ILOs 1, 3, 4). State the null and alternative hypotheses for the risk ratio and the odds ratio, noting the scale their null value sits on. Using the same method as the `arm`/`improved` example above, check with `levels()` which category of `gender` is already the reference, then use `relevel()` to set it explicitly (even if it turns out to already be the one you want). Obtain a risk ratio and an odds ratio, each with a 95% CI, for improvement in men compared with women. Reconcile the software output with a hand calculation from the same 2×2 table, and write a two-sentence results statement.

**Exercise 6** (ILO 2). Using `power.t.test()`, find the sample size per group needed to detect a difference in `chol` of 0.3 mmol/L, and again for a difference of 0.8 mmol/L, both at `power = 0.8` and `sig.level = 0.05`. Write one sentence explaining why the two required sample sizes differ so much.

## Writing exercise

Using the `arm`/`improved` result from earlier in this chapter, write two sentences for `strep_tb`'s primary outcome. State the proportion improved in each arm, and report the risk ratio or the odds ratio (your choice), its 95% CI, and its p-value, together.

::: {.callout-note collapse='true'}
## Solutions

**Exercise 1**

**Null hypothesis.** In the population, mean sedentary time is 6 hours/day (t-test); median sedentary time is 6 hours/day (`wilcox.test()`).

**Alternative hypothesis.** The mean, or the median, is not 6 hours/day.

```{r}
#| label: sol-ex1-onesample
t.test(nhanes$sed, mu = 6)
wilcox.test(nhanes$sed, mu = 6)
```

R's CI, 6.45 to 6.68 hours/day, is for the mean itself; subtracting the 6-hour benchmark from both ends gives the effect size: mean sedentary time is about 6.56 hours/day, 0.56 hours/day above 6 hours/day (95% CI: 0.45 to 0.68 hours/day above, p < 0.001). The Wilcoxon test is significant too (p < 0.001), but it's testing a different quantity, so its effect size is a median, not a mean: sample median 5 hours/day (IQR: 3 to 8 hours/day), *below* the 6-hour benchmark. The mean and the median disagree on which side of 6 hours 'typical' sedentary time falls. Sedentary time is right-skewed: a small number of participants report very high values, which pulls the mean up above the median, and above the benchmark, while the middle of the distribution, where most participants sit, remains below it.

**Exercise 2**

**Null hypothesis.** In the population of participants in this more sedentary subgroup, the mean within-person difference between `met_w` and `met_r` is zero.

**Alternative hypothesis.** The mean within-person difference is not zero.

```{r}
#| label: sol-ex2
sed_above_median <- nhanes |>
  filter(!is.na(sed), sed > median(sed, na.rm = TRUE))

tt_subgroup <- t.test(sed_above_median$met_w, sed_above_median$met_r, paired = TRUE)
tt_subgroup
wilcox.test(sed_above_median$met_w, sed_above_median$met_r, paired = TRUE)
```

Both tests still reject the null hypothesis in this more sedentary subgroup, though the mean difference is much smaller than in the whole sample (about `r round(tt_subgroup$estimate, 2)` MET-hours/day here, 95% CI: `r round(tt_subgroup$conf.int[1], 2)` to `r round(tt_subgroup$conf.int[2], 2)` MET-hours/day, p < 0.001, versus about 3.07 MET-hours/day overall). `met_w` and `met_r` remain two measurements taken on the same person at the same time, so the comparison is paired regardless of filtering.

**Exercise 3**

**Null hypothesis.** Mean age is the same in men and women in the population.

**Alternative hypothesis.** Mean age differs between men and women in the population.

```{r}
#| label: sol-ex3
ggplot(nhanes |> filter(!is.na(gender)), aes(x = gender, y = age)) +
  geom_boxplot() +
  labs(x = 'Gender', y = 'Age (years)')

t.test(age ~ gender, data = nhanes |> filter(!is.na(gender)))
```

This time R's CI is already positive, and that's informative: the difference is still men minus women, so a positive CI means the first group is the higher one. Men in this sample were on average about 0.7 years older than women (95% CI: 0.4 to 1.1 years, p < 0.001), and no signs need reversing to say so. The difference is statistically significant but tiny relative to a population averaging around 47 years old. With a sample this large, tens of thousands of participants, even a very small true difference will produce a small p-value. Statistical significance here is not the same as a practically important difference, and it doesn't contradict Chapter 3's impression that age looked similar across the two groups.

**Exercise 4**

**Null hypothesis.** In the population, improvement is not associated with baseline cavitation.

**Alternative hypothesis.** Improvement depends on baseline cavitation.

```{r}
#| label: sol-ex4
tab_cavitation <- table(strep_tb$baseline_cavitation, strep_tb$improved)
tab_cavitation
chisq.test(tab_cavitation, correct = FALSE)
chisq.test(tab_cavitation, correct = FALSE)$expected
```

All expected counts are comfortably above 5, the smallest being just under 22, so `correct = FALSE` is the right call here, for the same reason it was for the `arm` table.

The chi-squared test gives a p-value of about 0.056. That's above the conventional 0.05 threshold, but only just, and it would be a mistake to read much into which side of it the number landed on. The proportions behind it: 62% improved among patients without baseline cavitation, against 44% among those with it, an 18 percentage point gap the test can't confirm at this sample size. The honest reading is the same either way: this is weak evidence of an association between baseline cavitation and improvement. A trial of 107 patients cannot settle a question this size, and the result is best reported as inconclusive.

**Exercise 5**

**Null hypothesis.** In the population, the risk ratio is 1 and the odds ratio is 1: no association between gender and improvement.

**Alternative hypothesis.** The risk ratio, and the odds ratio, differ from 1. Both nulls sit at 1, not 0, since these are ratios instead of differences.

```{r}
#| label: sol-ex5
levels(strep_tb$gender)

gender_f_ref <- relevel(strep_tb$gender, ref = 'F')
tab_gender <- table(gender_f_ref, strep_tb$improved)
tab_gender

riskratio(tab_gender)
oddsratio(tab_gender, method = 'wald')
```

`levels()` shows that `'F'` is already listed first, so `relevel()` here confirms that choice explicitly. By hand, from the table: 27 of 48 men improved (risk = 0.5625) and 28 of 59 women improved (risk = 0.4746), giving a risk ratio of 0.5625 / 0.4746 = 1.19, and an odds ratio of (27 × 31) / (21 × 28) = 1.42. Both match the software output. The proportion improved was slightly higher in men than in women (risk ratio 1.19, 95% CI: 0.82 to 1.71; odds ratio 1.42, 95% CI: 0.66 to 3.06, chi-squared p = 0.37). But both CIs include the null value of 1, so this trial gives no good evidence of a difference in improvement by gender. `oddsratio()` prints three p-values, as it did for the `arm` table; the chi-squared one is quoted here because that's the one that reconciles with `chisq.test()`.

**Exercise 6**

```{r}
#| label: sol-ex6
power.t.test(delta = 0.3, sd = chol_sd, sig.level = 0.05, power = 0.8)
power.t.test(delta = 0.8, sd = chol_sd, sig.level = 0.05, power = 0.8)
```

Detecting the smaller difference (0.3 mmol/L) needs a much larger sample than detecting the larger difference (0.8 mmol/L). A smaller effect is harder to distinguish from chance variation, so we need more participants for the same 80% chance of detecting it.

**Writing exercise (sample answer)**

Among patients receiving streptomycin, 69% (38 of 55) showed radiological improvement at six months, compared with 33% (17 of 52) among those receiving the control treatment. This corresponds to a risk ratio of 2.11 (95% CI: 1.38 to 3.24, p < 0.001): about twice the proportion improved with streptomycin compared with control in this trial.
:::
