# Multiple linear and logistic regression (Week 11) {#sec-11-adjustment}

A confounder is a variable that lies behind an association and distorts it, and Chapter 6's single-predictor model has no way to hold one still. Baseline cardiovascular disease in our data is measured at the same visit as the exposure, so it's a real instance of the problem: nothing in the data says whether it sits upstream of sedentary behaviour or downstream of it. Whether it belongs in the model as a confounder to adjust for, or should stay out because it's a mediator, is a judgement the study design can't make for us, and this chapter fits the model both ways and reports both.

Getting there means extending linear regression to hold several predictors at once, so an adjusted coefficient can sit next to the crude one and the gap between them read as an estimate of confounding bias. The same extension handles an outcome that isn't continuous at all: whether someone died within ten years of their baseline visit becomes the input to a logistic regression, with odds ratios read straight off its exponentiated coefficients.

## Chapter intended learning outcomes

By the end of this chapter you will be able to:

1. Fit a multiple linear regression using `lm()`, including categorical predictors with a stated reference level
2. Compare crude and adjusted estimates and quantify the change
3. Fit a logistic regression with `glm(family = binomial)`, and interpret an exponentiated coefficient as an adjusted odds ratio with a confidence interval
4. Compare nested models as a sensitivity analysis, and appraise what each assumes about `blm.cvd`
5. Write an interpretation of an adjusted association in an appropriate academic style

## Setup

As in Chapters 3 to 6, we read the data with `read_csv()` and turn the six coded variables into factors against the dictionary.

This chapter adds one new package underneath the scenes: `gtsummary`'s `tbl_regression()`, which we use for the first time here, depends on `broom.helpers`. If `broom.helpers` isn't already on your machine, install it the usual way, in the console, not in the script. We don't `library()` it directly; `gtsummary` finds it automatically once it's installed.

```{r}
#| label: setup
#| message: false

library(readr)
library(dplyr)
library(ggplot2)
library(gtsummary)
library(here)
library(broom)

nhanes <- read_csv(here('data', 'mph_nhanes_20241107.csv'))
dictionary <- read_csv(here('data', 'mph_nhanes_20241107_dictionary.csv'))
```

```{r}
#| label: setup-hidden
#| include: false
options(width = 76)
source(here('R', 'gtsummary-knitprint.R'))
```

```{r}
#| label: recode

nhanes <- nhanes |>
  mutate(
    gender = factor(gender,
      levels = c(1, 2),
      labels = c('Male', 'Female')),
    eth = factor(eth,
      levels = c(1, 2, 3, 4, 5),
      labels = c('Mexican American', 'Other Hispanic', 'White', 'Black', 'Others')),
    edu = factor(edu,
      levels = c(1, 2, 3),
      labels = c('Less than high school', 'High school graduate', 'Above high school')),
    smoking = factor(smoking,
      levels = c('Never', 'Previous', 'Current')),
    phq.c = factor(phq.c,
      levels = c('0 Normal', '1 Mild symptoms', '2 Possible depression'),
      labels = c('Normal', 'Mild symptoms', 'Possible depression'),
      ordered = TRUE),
    blm.cvd = factor(blm.cvd,
      levels = c(0, 1),
      labels = c('No', 'Yes'))
  )
```

This is the same recode we introduced in Chapter 3 and repeated in Chapters 4 to 6. See Chapter 3 for the reasoning behind each variable.

## Opening exercise: building a ten-year mortality outcome

`nhanes_mph` records whether each participant died during follow-up (`dth`: 1=died; 0=alive) and how long they were followed for (`fu_month`, in months). Neither column on its own answers the question this chapter needs: did a participant die *within ten years*? 'Died' isn't a question we can answer until we say 'by when.'

Three groups exist in this data:

- Participants who died, and whose death happened within 120 months (10 years) of their follow-up starting. These count as an event: `1`.
- Participants known to be alive at 120 months. This includes everyone followed at least that long without dying, but it also includes anyone who died *after* 120 months: surviving to month 150 and then dying still means they were alive at month 120, so this group counts as a non-event: `0`, regardless of whether death eventually happened later on.
- Participants who did not die, but whose recorded follow-up ended before 120 months. These cannot be classified either way. They might have died the month after their follow-up ended, or lived another thirty years. Calling them 'survivors' would be a guess, so they are excluded rather than coded as `0`.

Two new tools here: `case_when()`, from `dplyr`, checks a series of conditions in order and returns the first matching value, which is exactly what a three-way rule needs. Each line pairs a condition on the left of the `~` with the value to return on the right. The last line is usually a catch-all: `TRUE` is a condition that is always met, so anything reaching it without matching a line above gets whatever that line returns. Order matters, because `case_when()` stops at the first condition a participant matches. The second tool is an argument rather than a function. `table()` normally leaves missing values out of its count, the way it did in Chapter 3. `useNA = 'always'` tells it to report them as their own category instead. That's what we want here, since the participants we exclude are the point, not a footnote.

::: {.callout-important appearance='simple'}
## Warm-up exercise

**Exercise.** Write the `mutate()` yourself. Using `case_when()`, build a new column called `dead_10yr` in `nhanes` that returns `1` for the first group above, `0` for the second, and `NA` for the third, then count the three groups with `table(nhanes$dead_10yr, useNA = 'always')`.

Two things to work out as you write it, which matter more than the syntax:

1. The rule for the second group needs no mention of `dth` at all. Why not? Think about who is in that group.
2. Once you have the counts, be ready to say why the third group can't be added to the `0`s and called survivors.

Have a go before opening the answer. Everything after this section runs on the column you're about to build, so build it yourself instead of reading someone else's version of it.
:::

::: {.callout-note collapse='true'}
## Answer: building `dead_10yr`

```{r}
#| label: dead-10yr
nhanes <- nhanes |>
  mutate(
    dead_10yr = case_when(
      dth == 1 & fu_month <= 120 ~ 1,
      fu_month >= 120 ~ 0,
      TRUE ~ NA
    )
  )

table(nhanes$dead_10yr, useNA = 'always')
```

The second condition is written as `fu_month >= 120` alone, with no condition on `dth`, precisely to catch both parts of that second group: `dth == 1` (died, but after month 120) and `dth == 0` (never died, followed at least that long) both belong in it. Anyone who was still being observed at month 120 was alive at month 120, and that is the whole question.

The catch-all comes last and returns `NA`. That's the third group: alive, but followed for less than ten years. Had it been written first, it would have swallowed everybody.
:::

```{r}
#| label: dead-10yr-counts
#| include: false
n_dead <- sum(nhanes$dead_10yr == 1, na.rm = TRUE)
n_alive <- sum(nhanes$dead_10yr == 0, na.rm = TRUE)
n_excluded <- sum(is.na(nhanes$dead_10yr))
```

`r format(n_dead, big.mark = ',')` participants died within the ten-year window, `r format(n_alive, big.mark = ',')` are known to have been alive at ten years (whether or not they later died), and `r format(n_excluded, big.mark = ',')` are excluded because their follow-up ended too early to classify them either way. That's a large number excluded, but it's a cost of a fixed-window outcome. We're deliberately not using information about *when* someone died, only whether it was inside or outside our chosen window. The `r format(n_dead, big.mark = ',')` events that remain are plenty to fit the regression models this chapter builds.

There is a family of methods, survival analysis, built precisely to use the follow-up time we've just thrown away and to keep the participants we've just excluded. Kaplan-Meier curves and Cox regression are its familiar tools. They're taught in Advanced Statistics.

::: {.callout-note}
## Try it yourself: a different window

Rebuild `dead_10yr` using a five-year window (60 months) instead of ten. Report the new event and exclusion counts, then explain in one sentence why a shorter window moves both numbers in the direction it does.
:::

## Extending the Chapter 6 model

This section adjusts for known confounders: age, gender, and education. Chapter 6 fitted total cholesterol (`chol`, in mmol/L) against sedentary time alone (`sed`, in hours/day), `chol ~ sed`, and found a small, non-significant slope: it adjusted for nothing else, so it couldn't rule out age, gender, and education masking the association. We now add them using `+` to an `lm()` formula.

**Null hypothesis.** In the population, the adjusted slope of cholesterol on `sed` is zero, once age, gender, and education are held constant.

**Alternative hypothesis.** The adjusted slope is not zero.

```{r}
#| label: lm-adjusted
chol_adj_complete <- nhanes |>
  filter(!is.na(sed), !is.na(chol), !is.na(age), !is.na(gender), !is.na(edu))

m_adj <- lm(chol ~ sed + age + gender + edu, data = chol_adj_complete)
summary(m_adj)
confint(m_adj)
```

The coefficient on `sed` is now what this model calls the *adjusted* association between sedentary time and cholesterol: the association after holding age, gender, and education constant. 'Adjustment' here means the model assumes the effect of an extra hour of sedentary time is the same regardless of a participant's age, gender, or education, and estimates that one shared slope from everybody at once.

`gender` and `edu` entered that model as factors, and each one contributes a row per non-reference level rather than one row for the whole variable. `levels()`, from base R, shows a factor's categories in order; the first one listed is the reference level, the category every other row is compared against, the same idea Chapter 6 covered in full, including what `relevel()` does and doesn't change. We don't interpret `gender`'s or `edu`'s own coefficients here the way we interpret `sed`'s: neither was the exposure this model was built to assess, so their adjustment sets weren't chosen with that in mind. More on exactly why, below.

## Crude versus adjusted: quantifying the change

Chapter 6's crude result and this chapter's adjusted one belong side by side. We refit the crude model here instead of reusing Chapter 6's, so both models run on `chol_adj_complete`, the same set of participants. Without aligning the samples, any difference between the two estimates could be attributed to either the adjustment or a difference in sample size, and we couldn't tell which. This is also why the crude figure below isn't identical to the one Chapter 6 reported.

```{r}
#| label: crude-model
m_crude <- lm(chol ~ sed, data = chol_adj_complete)
tidy(m_crude, conf.int = TRUE) |> filter(term == 'sed')
tidy(m_adj, conf.int = TRUE) |> filter(term == 'sed')
```

```{r}
#| label: crude-adj-numbers
#| include: false
crude_sed <- tidy(m_crude, conf.int = TRUE) |> filter(term == 'sed')
adj_sed <- tidy(m_adj, conf.int = TRUE) |> filter(term == 'sed')

# sprintf() rather than round(): these slopes are small enough that round()
# hands knitr a number in scientific notation, which renders as 7.4^{-4}.
# dp3 keeps odds ratios at a fixed three decimals, so 1.020 doesn't print as
# 1.02 alongside 1.019 in a sentence comparing the two.
dp5 <- function(x) sprintf('%.5f', x)
dp3 <- function(x) sprintf('%.3f', x)
```

Crude, the slope on `sed` is about `r dp5(crude_sed$estimate)` mmol/L per hour/day (95% CI `r dp5(crude_sed$conf.low)` to `r dp5(crude_sed$conf.high)` mmol/L per hour/day, p = `r round(crude_sed$p.value, 3)`). Adjusted for age, gender, and education, it moves to about `r dp5(adj_sed$estimate)` mmol/L per hour/day (95% CI `r dp5(adj_sed$conf.low)` to `r dp5(adj_sed$conf.high)` mmol/L per hour/day, p = `r round(adj_sed$p.value, 3)`).

Look at the estimate first, not the p-value. The adjusted slope is about `r round(adj_sed$estimate / crude_sed$estimate, 1)` times the size of the crude one: holding age, gender, and education constant pulls the association between sedentary time and cholesterol substantially further from zero. That movement is what confounding looks like. Age and gender were masking part of a real association, and the model that ignores them understates it. The p-value crossing 0.05 is a consequence of the estimate moving away from zero, not a separate finding.

A caution about reading the other rows of this table: `age`'s coefficient here is adjusted for `sed` too, the same as `sed`'s own coefficient is. If sedentary time sits on the path from age to cholesterol, part of age's true association runs through it, and `age`'s row only captures the part that doesn't. Treating every coefficient in a regression table as though it carried the same interpretation as the one we set out to estimate is called the *Table 2 fallacy*. `sed` is the exposure this model was built to assess, and its adjustment set was chosen with that in mind; `age`, `gender`, and `edu` weren't, so their coefficients shouldn't be read as if they were each other's crude-versus-adjusted comparisons in miniature. This is the same logic the sensitivity analysis later in this chapter applies to `blm.cvd`, just stated as a general caution rather than worked through on one variable.

## Diagnostics for the multiple regression model

The same diagnostic plots Chapter 6 used still apply here, checked against the adjusted model rather than the simple one:

```{r}
#| label: diagnostics-adjusted
adj_aug <- augment(m_adj)

ggplot(adj_aug, aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.1) +
  geom_hline(yintercept = 0) +
  labs(x = 'Fitted value (mmol/L)', y = 'Residual (mmol/L)')

ggplot(adj_aug, aes(sample = .std.resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(x = 'Theoretical quantiles', y = 'Standardised residuals')
```

The same pattern Chapter 6 found reappears: a fairly even scatter around zero for most of the range, with a small group of participants with unusually high cholesterol pulling the upper tail away from the reference line in the QQ plot.

## When the outcome is binary: logistic regression

The question for the rest of this chapter: is sedentary time associated with ten-year mortality, and does that association hold up once potential confounders are adjusted for? `chol` is continuous, so `lm()` was the right tool. But `dead_10yr` is 0 or 1, and we can't model it the same way. An `lm()` produces a straight line, and a straight line fitted to a 0/1 outcome can predict values below 0 or above 1, which aren't valid probabilities. `glm()`, base R's function for generalised linear models (a family of regression models taught further in Advanced Statistics), fixes this by modelling the log odds of the outcome instead of the outcome directly.

We build the complete-case sample first and fit both models on it, for the same reason the linear models above shared one sample: the crude and adjusted estimates should differ because of the adjustment, not because they were calculated on different people.

```{r}
#| label: dead-complete
dead_complete <- nhanes |>
  filter(!is.na(dead_10yr), !is.na(sed), !is.na(age), !is.na(gender), !is.na(edu))
```

Fitting one looks almost identical to fitting an `lm()`. The formula is the same shape, and the only addition is `family = binomial`, which tells `glm()` the outcome is binary.

**Null hypothesis.** In the population, sedentary time carries no association with ten-year mortality: on the odds ratio scale, met below, an odds ratio of 1.

**Alternative hypothesis.** Sedentary time is associated with ten-year mortality, an odds ratio away from 1.

```{r}
#| label: glm-crude
m_dead_crude <- glm(dead_10yr ~ sed, data = dead_complete, family = binomial)
summary(m_dead_crude)
```

```{r}
#| label: glm-adjusted
m_dead_adj <- glm(dead_10yr ~ sed + age + gender + edu,
  data = dead_complete, family = binomial)
summary(m_dead_adj)
```

We keep this adjustment set deliberately small. `dead_complete` holds `r format(nrow(dead_complete), big.mark = ',')` participants, of whom `r format(sum(dead_complete$dead_10yr), big.mark = ',')` died within ten years. That's fewer events than the `r format(sum(nhanes$dead_10yr, na.rm = TRUE), big.mark = ',')` we counted at the start of the chapter, because requiring a recorded value for every predictor drops participants of its own. Those `r format(sum(dead_complete$dead_10yr), big.mark = ',')` events are what the model has to work with, and there isn't room to add many more predictors before it runs short of events per predictor.

## Odds ratios: exponentiating the coefficients

A logistic regression's coefficients are on the log-odds scale, which isn't how the epidemiology half reported odds ratios. `exp()`, base R's exponential function, undoes the log and converts a coefficient back to an odds ratio. `coef()` and `confint()` pull the coefficients and their confidence intervals the same way they did for `lm()`:

```{r}
#| label: or-crude
exp(coef(m_dead_crude))
exp(confint(m_dead_crude))
```

`confint()` printed `Waiting for profiling to be done...` before the answer. That's a message, not a warning, and it marks a real difference between `glm()` and `lm()`.

For a linear model, `confint()` builds each CI the way Chapter 4 built one by hand: an estimate, plus or minus a multiplier times a standard error. For a logistic model that shortcut is only an approximation, so `confint()` does something better instead. It works along each coefficient, refitting the model repeatedly to find the range of values the data are consistent with. That search is the 'profiling', and the message is `glm()` saying it's running. On this dataset it takes a second or two. On a much larger model it can take noticeably longer, which is the only reason the message exists at all.

```{r}
#| label: or-adjusted
exp(coef(m_dead_adj))
exp(confint(m_dead_adj))
```

The same message appears again above, for the same reason.

`broom::tidy()` does the same conversion in one step with `exponentiate = TRUE`:

```{r}
#| label: tidy-exponentiate
tidy(m_dead_adj, exponentiate = TRUE, conf.int = TRUE) |> filter(term == 'sed')
```

```{r}
#| label: or-numbers
#| include: false
or_crude <- tidy(m_dead_crude, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term == 'sed')
or_adj <- tidy(m_dead_adj, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term == 'sed')
```

Crude, each additional hour/day of sedentary time is associated with an odds ratio of about `r dp3(or_crude$estimate)` for ten-year mortality (95% CI: `r dp3(or_crude$conf.low)` to `r dp3(or_crude$conf.high)`, p < 0.001). Adjusted for age, gender, and education, the odds ratio attenuates to about `r dp3(or_adj$estimate)` (95% CI: `r dp3(or_adj$conf.low)` to `r dp3(or_adj$conf.high)`, p < 0.001). Both CIs sit entirely above 1, the null value for an odds ratio, so both are statistically significant. But the adjusted estimate is noticeably smaller than the crude one. Part of the crude association was really age. Older participants are both more sedentary and more likely to die within ten years, for reasons that have nothing to do with sedentary behaviour itself.

Both odds ratios sit close to 1, so confounding matters less here than it will in Exercise 2, where the same mechanism, age, moves the estimate by a factor of about six instead of a couple of percent. Do that exercise before deciding how seriously to take adjustment.

## Regression tables with `gtsummary`

`tbl_regression()`, from `gtsummary`, builds a publication-style table directly from a fitted model, the regression equivalent of `tbl_summary()`. For the linear model, no extra argument is needed:

```{r}
#| label: tbl-regression-linear
tbl_regression(m_adj)
```

For the logistic model, `exponentiate = TRUE` reports odds ratios instead of log-odds coefficients:

```{r}
#| label: tbl-regression-logistic
tbl_regression(m_dead_adj, exponentiate = TRUE)
```

`tbl_merge()`, also from `gtsummary`, places two tables side by side under named headers, which is exactly what a crude-vs-adjusted comparison needs:

```{r}
#| label: tbl-merge-crude-adjusted
tbl_crude <- tbl_regression(m_dead_crude, exponentiate = TRUE, include = 1)
tbl_adj <- tbl_regression(m_dead_adj, exponentiate = TRUE, include = 1)

tbl_merge(list(tbl_crude, tbl_adj),
  tab_spanner = c('Crude', 'Adjusted'))
```

`include = 1` keeps only the `sed` row, the estimate we're interested in, and drops the adjustment variables' own rows. That's fine here: the Table 2 fallacy above is why we weren't going to interpret those rows anyway. The table still needs a caption or footnote naming which variables were adjusted for, since that information doesn't disappear just because the rows do.

## The sensitivity analysis: is `blm.cvd` a confounder or a mediator?

The seminar raised a genuine difficulty with `blm.cvd`, baseline cardiovascular disease. The two models below are **nested**: one model's predictors, `sed + age + gender + edu`, are a subset of the other's, which adds only `blm.cvd`. Comparing a nested pair like this isolates the effect of the one variable that differs between them. It's plausible as a confounder. Existing heart disease could make someone both less physically active and more likely to die, and adjusting for it would then remove bias. It's equally plausible as a mediator. Sedentary behaviour could contribute to cardiovascular disease, which then contributes to death, and adjusting for it would then remove part of the total effect we're trying to estimate. Baseline CVD and sedentary time were both measured at the same study visit, so nothing in the data itself establishes which came first. That's not a gap in this analysis. It's a limit on what an observational snapshot can ever tell us. The honest response is to report the estimate both ways instead of picking one.

```{r}
#| label: sensitivity-cvd
dead_cvd_complete <- dead_complete |> filter(!is.na(blm.cvd))

m_dead_without <- glm(dead_10yr ~ sed + age + gender + edu,
  data = dead_cvd_complete, family = binomial)
m_dead_with <- glm(dead_10yr ~ sed + age + gender + edu + blm.cvd,
  data = dead_cvd_complete, family = binomial)

tbl_without <- tbl_regression(m_dead_without, exponentiate = TRUE, include = 1)
tbl_with <- tbl_regression(m_dead_with, exponentiate = TRUE, include = 1)

tbl_merge(list(tbl_without, tbl_with),
  tab_spanner = c('Without baseline CVD', 'With baseline CVD'))
```

```{r}
#| label: sensitivity-numbers
#| include: false
or_without <- tidy(m_dead_without, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term == 'sed')
or_with <- tidy(m_dead_with, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term == 'sed')
```

The odds ratio for `sed` moves only a little between the two models: `r dp3(or_without$estimate)` (95% CI `r dp3(or_without$conf.low)` to `r dp3(or_without$conf.high)`, p < 0.001) without `blm.cvd`, and `r dp3(or_with$estimate)` (95% CI `r dp3(or_with$conf.low)` to `r dp3(or_with$conf.high)`, p < 0.001) with it. `blm.cvd` itself is a strong predictor of ten-year mortality, but adjusting for it changes almost nothing about the sedentary time estimate. Whichever role `blm.cvd` plays, confounder or mediator, the conclusion about sedentary time and mortality in this dataset doesn't depend much on which one it is. 

## Exercises

**Exercise 1** (ILOs 1, 2). State the null and alternative hypotheses for the adjusted slope. Fit `chol ~ mpa_w + age + gender + edu`, where `mpa_w` is moderate physical activity at work in hours/day, and compare it with the crude model `chol ~ mpa_w`. Report both slopes for `mpa_w` with their 95% CIs and p-values, and write one sentence describing how the adjusted estimate differs from the crude one, including its direction.

**Exercise 2** (ILOs 3, 5). State the null and alternative hypotheses for the odds ratio, noting the scale its null value sits on. Fit `dead_10yr ~ vpa_t + age + gender + edu`, where `vpa_t` is vigorous physical activity overall in hours/day, and compare it with the crude model `dead_10yr ~ vpa_t`. Exponentiate both models' coefficients to obtain odds ratios with CIs, and build a `tbl_regression()` table for the adjusted model. Write one sentence interpreting the adjusted odds ratio for `vpa_t` in context.

**Exercise 3** (ILO 4). Using the adjusted model from Exercise 2, fit a second version that also adjusts for `blm.cvd`. Combine both models into one table with `tbl_merge()`, and write one sentence comparing the two odds ratios for `vpa_t` and one sentence stating what each model assumes about `blm.cvd`.

## Writing exercise

Write two short paragraphs, as they would appear in a results section.

In the first, report the adjusted association between sedentary time and cholesterol from this chapter's linear model, and the adjusted odds ratio for sedentary time and ten-year mortality from the logistic model, each with its 95% CI and p-value. Add one sentence on what this analysis cannot establish: an observational adjustment reduces confounding by the variables included, but it cannot rule out confounding by variables that were never measured.

In the second, separate paragraph, report both odds ratios for `sed` from the sensitivity analysis, with and without `blm.cvd`, and state plainly what each model assumes about baseline CVD's role and why the data available here cannot decide between those assumptions.

::: {.callout-note collapse='true'}
## Solutions

**Exercise 1**

**Null hypothesis.** In the population, the adjusted slope of cholesterol on `mpa_w` is zero, once age, gender, and education are held constant.

**Alternative hypothesis.** The adjusted slope is not zero.

```{r}
#| label: sol-ex1
mpa_w_complete <- nhanes |>
  filter(!is.na(mpa_w), !is.na(chol), !is.na(age), !is.na(gender), !is.na(edu))

m_mpa_w_crude <- lm(chol ~ mpa_w, data = mpa_w_complete)
m_mpa_w_adj <- lm(chol ~ mpa_w + age + gender + edu, data = mpa_w_complete)

tidy(m_mpa_w_crude, conf.int = TRUE) |> filter(term == 'mpa_w')
tidy(m_mpa_w_adj, conf.int = TRUE) |> filter(term == 'mpa_w')
```

```{r}
#| label: sol-ex1-numbers
#| include: false
crude_mpa <- tidy(m_mpa_w_crude, conf.int = TRUE) |> filter(term == 'mpa_w')
adj_mpa <- tidy(m_mpa_w_adj, conf.int = TRUE) |> filter(term == 'mpa_w')
```

Crude, the slope on `mpa_w` is about `r dp5(crude_mpa$estimate)` mmol/L per hour/day (95% CI: `r dp5(crude_mpa$conf.low)` to `r dp5(crude_mpa$conf.high)` mmol/L per hour/day, p = `r round(crude_mpa$p.value, 3)`). The CI sits almost symmetrically either side of zero, so on its own this says very little: these data are consistent with no association between moderate activity at work and cholesterol, and they don't rule one out either.

Adjusted for age, gender, and education, the slope changes sign and becomes non-zero, at about `r dp5(adj_mpa$estimate)` mmol/L per hour/day (95% CI: `r dp5(adj_mpa$conf.low)` to `r dp5(adj_mpa$conf.high)` mmol/L per hour/day, p = `r round(adj_mpa$p.value, 3)`). Once those three are held constant, more moderate activity at work is associated with slightly *higher* cholesterol in this sample. That's a reminder that adjustment can change a result's direction as well as its size, and that a crude estimate near zero can hide an association instead of ruling one out. Both models are fitted on `mpa_w_complete`, the same participants, so the difference between them is the adjustment and nothing else.

```{r}
#| label: sol-ex1-levels
levels(mpa_w_complete$edu)
```

`'Less than high school'` is listed first, so that's the reference this model used: both `edu` rows in the adjusted output above are differences from that group. We don't read them further than that here, for the same reason the chapter gave: `edu` wasn't the exposure this model was built to estimate.

**Exercise 2**

**Null hypothesis.** In the population, the adjusted odds ratio for `vpa_t` is 1: no association between vigorous activity overall and ten-year mortality, once age, gender, and education are held constant.

**Alternative hypothesis.** The adjusted odds ratio is not 1. The null sits at 1, not 0, since this is a ratio.

```{r}
#| label: sol-ex2
vpa_t_dead_complete <- nhanes |>
  filter(!is.na(vpa_t), !is.na(dead_10yr), !is.na(age), !is.na(gender), !is.na(edu))

m_vpa_t_crude <- glm(dead_10yr ~ vpa_t, data = vpa_t_dead_complete, family = binomial)
m_vpa_t_adj <- glm(dead_10yr ~ vpa_t + age + gender + edu,
  data = vpa_t_dead_complete, family = binomial)

tidy(m_vpa_t_crude, exponentiate = TRUE, conf.int = TRUE) |> filter(term == 'vpa_t')
tidy(m_vpa_t_adj, exponentiate = TRUE, conf.int = TRUE) |> filter(term == 'vpa_t')

tbl_regression(m_vpa_t_adj, exponentiate = TRUE)
```

```{r}
#| label: sol-ex2-numbers
#| include: false
or_vpa_crude <- tidy(m_vpa_t_crude, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term == 'vpa_t')
or_vpa_adj <- tidy(m_vpa_t_adj, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term == 'vpa_t')
```

Adjusted for age, gender, and education, each additional hour/day of vigorous activity overall is associated with an odds ratio of about `r dp3(or_vpa_adj$estimate)` for ten-year mortality (95% CI: `r dp3(or_vpa_adj$conf.low)` to `r dp3(or_vpa_adj$conf.high)`, p < 0.001): participants reporting more vigorous activity had lower odds of dying within ten years, holding age, gender, and education constant.

Now compare that with the crude model. Crude, the odds ratio is `r dp3(or_vpa_crude$estimate)`, which would say that an extra hour a day of vigorous activity cuts the odds of dying by something like `r round((1 - or_vpa_crude$estimate) * 100)`%. Adjusted, it's `r dp3(or_vpa_adj$estimate)`: still a real association, but noticeably weaker. Part of that gap is age. Older participants do less vigorous activity overall, and are more likely to die within ten years, for reasons that have nothing to do with the activity itself. So a crude comparison of the active against the inactive is partly a comparison of the young against the old. This is the same confounding story the chapter told with `sed`, moving the estimate the same way, toward the null: there the adjusted odds ratio moved from `r dp3(or_crude$estimate)` to `r dp3(or_adj$estimate)`, closer to 1; here it moves from `r dp3(or_vpa_crude$estimate)` to `r dp3(or_vpa_adj$estimate)`, also closer to 1, from further below it. Reporting the crude figure on its own would overstate what vigorous activity does.

**Exercise 3**

```{r}
#| label: sol-ex3
vpa_t_cvd_complete <- vpa_t_dead_complete |> filter(!is.na(blm.cvd))

m_vpa_t_without <- glm(dead_10yr ~ vpa_t + age + gender + edu,
  data = vpa_t_cvd_complete, family = binomial)
m_vpa_t_with <- glm(dead_10yr ~ vpa_t + age + gender + edu + blm.cvd,
  data = vpa_t_cvd_complete, family = binomial)

tbl_vpa_without <- tbl_regression(m_vpa_t_without, exponentiate = TRUE)
tbl_vpa_with <- tbl_regression(m_vpa_t_with, exponentiate = TRUE)

tbl_merge(list(tbl_vpa_without, tbl_vpa_with),
  tab_spanner = c('Without baseline CVD', 'With baseline CVD'), quiet = TRUE)
```

```{r}
#| label: sol-ex3-numbers
#| include: false
or_vpa_without <- tidy(m_vpa_t_without, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term == 'vpa_t')
or_vpa_with <- tidy(m_vpa_t_with, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term == 'vpa_t')
```

The odds ratio for `vpa_t` moves a little between the two models: `r dp3(or_vpa_without$estimate)` (95% CI `r dp3(or_vpa_without$conf.low)` to `r dp3(or_vpa_without$conf.high)`, p < 0.001) without `blm.cvd`, and `r dp3(or_vpa_with$estimate)` (95% CI `r dp3(or_vpa_with$conf.low)` to `r dp3(or_vpa_with$conf.high)`, p < 0.001) with it. That's a smaller shift than the main adjustment step but still a real one. Without `blm.cvd`, the model assumes that any relationship between baseline CVD and mortality operates entirely through the predictors already in the model; with it, the model assumes `blm.cvd` is a separate cause of death, worth adjusting for directly. Both are defensible readings of a single measurement taken at one study visit, and nothing in this dataset can say which one is correct.

**Writing exercise (sample answer)**

Sedentary time was associated with total cholesterol after adjustment for age, gender, and education (slope `r dp5(adj_sed$estimate)` mmol/L per hour/day, 95% CI `r dp5(adj_sed$conf.low)` to `r dp5(adj_sed$conf.high)` mmol/L per hour/day, p = `r round(adj_sed$p.value, 3)`), and with higher odds of ten-year all-cause mortality (adjusted odds ratio `r dp3(or_adj$estimate)`, 95% CI `r dp3(or_adj$conf.low)` to `r dp3(or_adj$conf.high)`, p < 0.001). Both estimates adjust only for the variables included in this model; they cannot rule out confounding by factors that were never measured, such as overall frailty or comorbidity not captured here.

In a sensitivity analysis, the adjusted odds ratio for sedentary time and ten-year mortality was `r dp3(or_without$estimate)` (95% CI `r dp3(or_without$conf.low)` to `r dp3(or_without$conf.high)`, p < 0.001) without baseline cardiovascular disease in the model, and `r dp3(or_with$estimate)` (95% CI `r dp3(or_with$conf.low)` to `r dp3(or_with$conf.high)`, p < 0.001) with it included. The first model assumes baseline CVD's influence on mortality runs entirely through the other adjustment variables; the second treats it as an independent contributor to mortality risk worth adjusting for directly. Because sedentary behaviour and baseline CVD were both measured at the same study visit, this dataset cannot establish which assumption is correct, and both estimates are reported for that reason.
:::
